Simple Methods to Acutely Measure Multiple Timing Metrics among Sexual Repertoire of Male Drosophila

Version 1

Abstract

Male Drosophila courtship behavior is a key model for studying temporal decision-making. While courtship index (CI) is widely used to quantify mating activity, other timing metrics like courtship latency, copulation latency (CL), and mating duration (MD) remain understudied. Traditional methods for quantifying these behaviors are often labor-intensive and prone to human error.

In this study, we present a protocol combining a modular chamber system and automated software (DrosoMating) to quantify 6 key timing metrics during male courtship. High-resolution video tracking enable precise analysis of courtship initiation, persistence, and mating dynamics under controlled conditions. Validation shows <0.05% error rates and 98–99% agreement with manual scoring for CL, CI, and MD. The protocol supports genetic and neural circuit manipulations, detecting subtle genetic, social, and environmental behavioral variations.

By minimizing manual effort and standardizing data collection, this approach facilitates scalable, reproducible studies on adaptive trade-offs, learning, and neural mechanisms in mating behavior. This method streamlines timing analysis in male courtship, offering reproducible metrics for behavioral genetics.”

Introduction

Time perception is critical for survival, enabling organisms to estimate intervals between events and execute adaptive behaviors. Male Drosophila courtship behavior, a genetically well-characterized model (Singh and Singh, 2016), provides a robust framework for studying temporal control in action sequences. A widely used metric to quantify mating activity is the ‘Courtship Index (CI)’, defined as the percentage of time a male engages in courtship behaviors (e.g., following, wing vibration, abdominal curling) toward a female during a standardized observation window. CI reflects both the intensity and persistence of courtship, serving as a sensitive readout for dissecting genetic, environmental, or experimental perturbations on mating behavior, as demonstrated in studies investigating genetic mutations, environmental perturbations, and neuronal manipulations (Benton, 2015; Greenspan and Ferveur, 2000a; Kim, 2009; Pavlou and Goodwin, 2013; Yamamoto and Koganezawa, 2013).

Despite extensive research on courtship displays (Chen et al., 2024a), other critical metrics remain poorly characterized, especially as their representative characteristics as timing behaviors. ‘Courtship latency’ (time until a male initiates courtship after detecting a female) may reflect motivational state or sensory processing deficits but is rarely quantified in terms of genetics (Eastwood and Burnet, 1977a). ‘Copulation Latency (CL)’ (time from courtship initiation to successful mating) and ‘Copulation (Mating) Duration (MD)’ (length of mating itself) are similarly underexplored, though both likely impact reproductive success. Prolonged MD enhances paternity assurance by maximizing sperm transfer and suppressing female remating, but it simultaneously increases predation risks and energy expenditure (Bretman et al., 2009). Conversely, reduced MD may decrease a male’s paternity share due to incomplete sperm transfer or female resistance, yet it reduces exposure to ecological threats (e.g., predators, rivals) and conserves metabolic resources (Lee et al., 2023a). These traits may be influenced by cryptic factors such as male age, rival presence, or environmental stressors (e.g., temperature), yet their genetic and neural bases are unclear. While studies focus on stereotyped courtship actions, these “temporal metrics” offer insights into decision-making, persistence, and adaptive trade-offs—areas ripe for investigation (Huang et al., 2024; Kim et al., 2012a, 2013a, 2016, 2024; Lee et al., 2022; D. Sun et al., 2024; Y. Sun et al., 2024; Wong et al., 2019; T. Zhang et al., 2024b, 2024a; X. Zhang et al., 2024).

Numerous software tools have been developed to measure male mating behavior in Drosophila, yet many laboratories still rely on manual methods. Traditional approaches, such as the “courtship wheel”, have been widely used to assess male courtship activity (Hall, 1994; Neckameyer and Bhatt, 2016; Vosshall, 2007). However, quantifying the full repertoire of mating behaviors metrics has historically been labor-intensive (Ejima and Griffith, 2007; Sokolowski, 2001). Recent advances in automation have addressed these challenges. For instance, Dankert et al., introduced a machine vision system to quantify male-female interactions, though its application has been more focused on studying male-male aggression (Dankert et al., 2009a). With the rapid development of machine learning and image processing, tools such as ‘Ctrax’ (Branson et al., 2009), ‘idTracker’ (Pérez-Escudero et al., 2014), Cadabra (Stern et al., 2015), ‘FlyTracker’ and MATLAB combined with ‘JAABA’ (Barwell et al., 2021; Kabra et al., 2013) have enabled precise quantification of male-female mating behavior. Recently, Chen et al., introduced an image-processing-based system that addresses this gap by automatically recognizing multiple courtship elements with high accuracy (>97%) and robust identity tracking even during complete fly overlap(Chen et al., 2024b). Despite these advancements, recent efforts to simplify courtship conditioning protocols (Gil-Martí et al., 2023a) still lack of certain timing-related behavioral repertoires after copulation. Consequently, there is a need for more streamlined protocols and software to accurately measure the timing aspects of male mating behavior in Drosophila.

Beyond Drosophila-specific tools, general behavioral video analysis paradigms have been developed to track the movement and behavior of various organisms, including fruit flies. These methods often focus on tracking the position of single or multiple flies (Iyengar et al., 2012; Qu et al., 2022; Simon et al., 2011) or using artificial markers (Gal et al., 2020). Additionally, programs like ‘ToxTrac’ (Rodriguez et al., 2017), ‘UMATracker’ (Yamanaka and Takeuchi, 2018), ‘FIMTrack’ (Risse et al., 2017, 2014), and ‘ilastik’ (Berg et al., 2019) have been employed to analyze body posture, behavior, location, and movement trajectories in diverse species, including flies, bees, mice, and zebrafish. However, these tools often fall short in capturing the nuanced courtship behaviors of Drosophila, as motion data alone may not sufficiently represent complex interactions.

To address these limitations, we have developed a specialized software called ‘DrosoMating’ suite for analyzing and quantifying timing behaviors within male Drosophila courtship. By establishing 6 key indicators related to male courtship, we provide a robust framework for quantifying male mating behavior metrics under controlled conditions. This approach significantly reduces the time and errors associated with manual identification. Through iterative optimization, our system achieves an error rate of less than 0.05% compared to manual scoring. We believe this toolset offers a novel and efficient method for studying timing in male courtship behavior, providing valuable data and theoretical insights while streamlining the analysis process.

Results

Video-Processing Solutions for Precise Drosophila Identification in Multi-Chamber Setups

Accurate identification of Drosophila individuals is essential for behavioral quantification given their small size. DrosoMating addresses this through three integrated video-processing features that enhance recognition fidelity in multi-chamber environments.

First, an effective recognition region function isolates individual wells as discrete analysis zones (Fig. 1A bottom), eliminating interference from adjacent chambers and background artifacts. Second, an image learning function adapts to phenotypic variability across strains. Operators designate three sample flies (Fig. 1B), enabling the algorithm to internalize strain-specific morphological signatures for improved detection. Third, a manual threshold adjustment function permits post-recognition optimization. Users visually validate automated identifications and dynamically refine sensitivity parameters (Fig. 1C), correcting false positives/negatives. Collectively, these features ensure robust tracking accuracy across diverse experimental setups and genetic backgrounds.

Usage of DrosoMating and CAD Parameters of the Chamber

(A) Typing input speed and input board numbers before selecting analyzing area (upper). Schematic representation of the region selection process (lower). Columns should be selected in a clockwise direction starting from the top left corner.. (B) Selection of reference flies for tracking accuracy. Any three reference flies were chosen for the analysis. (C) Use the X-axis, Y-axis, and threshold regulator to adjust the selection area to fit each well. Click “Enter” to run the program. Output file will be in the video folder: [Video Name]_output_[board numbers].csv (D) CAD exploded 3D view of the chamber instruction. (E) CAD parametric data of the chamber (dimensions in mm). (F) 3D diagram of the completed chamber

Development and Validation of DrosoMating for Automated Drosophila Mating Behavior Analysis

To address the labor-intensive nature, high error rates, and limitations in quantifying complex behavioral metrics during manual observation of Drosophila mating, we developed the software DrosoMating. This tool automates analysis of mating videos recorded in a customized chamber. DrosoMating processes videos to generate a CSV file containing 6 key behavioral metrics: courtship duration, MD, CI, courtship start time, mating start time, and mating end time. While these metrics are individually straightforward, they can be combined to quantify diverse aspects of mating behavior metrics.

DrosoMating enables quantification of specific mating behaviors based on research needs. We focused on three core indices:

Courtship Index (CI): Indicates male courtship investment per unit time. Higher CI values (reflecting longer durations of behaviors like wing vibration or female chasing) suggest stronger male courtship motivation or higher female attractiveness (Greenspan and Ferveur, 2000b).

Copulation Latency (CL): Measures the time from courtship initiation to copulation start. Shorter CL indicates efficient male courtship or high female receptivity. Longer CL may reflect male courtship deficits (Eastwood and Burnet, 1977b), intense female rejection, or environmental stressors (Greenspan and Ferveur, 2000b; Singh and Singh, 2014).

Mating Duration (MD): Represents the copulation period length. Longer MD may enhance sperm competition but increases predation risk and energy costs, illustrating an evolutionary trade-off between reproduction and survival (Greenspan and Ferveur, 2000b).

Together, CI, CL, and MD provide a robust framework for characterizing Drosophila behavioral states during mating. Accurate quantification of these metrics is essential.

Because overlapping behaviors occur frequently during courtship and mating. When flies overlap for ≤15 frames, centroid velocity continuity is used to assign identity; if overlap >15 frames, the region is flagged ‘occluded’ and excluded from CI calculation.

We validated DrosoMating by comparing its output for CI, CL, and MD against values from repeated, corrected manual measurements. Differences between software-generated data and manual data were consistently within 10 seconds for all three metrics, with no statistically significant differences (Fig. 2A-C). Manual quantification of these metrics requires several hours of effort and is prone to substantial human error. In contrast, DrosoMating automates the analysis. After a few minutes of initial setup and configuration, the software generates precise behavioral metrics from mating videos.

Comparison of the Accuracy of DrosoMating with Manual Scoring

(A) Copulation latency (CL) of Canton-S males was analyzed using DrosoMating and compared with manual recordings. DBMs represent the ‘difference between means’ for the evaluation of estimation statistics. Asterisks represent significant differences, as revealed by the Student’s t test (* p<0.05, ** p<0.01, *** p<0.001). See the Background for a detailed description of the CL assays used in this study. (B) Courtship index (CI) of Canton-S was analyzed by DrosoMating and manual recording. CI was measured by [courtship time/ (mating time-courtship time) *100%]. For detailed methods, see the BACKGROUND for a detailed description of the CI assays used in this study. (C) Mating Duration (MD) of Canton-S was analyzed by DrosoMating and manual recording. See the BACKGROUND for a detailed description of the MD assays used in this study.

Validation of DrosoMating Across Diverse Drosophila Strains and Behavioral Paradigms

DrosoMating was adapted to quantify mating behaviors metrics across diverse Drosophila strains and experimental conditions. To evaluate its universality and accuracy, mating interactions were analyzed between virgin females and males of four strains: Canton-S (wild-type), Oregon-R (wild-type), w1118(white eye mutant), and y1 (yellow body color mutant). These lines have been extensively employed as standard inbred strains in research contexts for a prolonged period. Wild-type strains (Canton-S and Oregon-R) exhibited significantly higher CI, CL, and MD compared to w1118 and y1 mutants (Fig. 3A–C). This quantitative assessment revealed markedly reduced activity in mutant males across all three behavioral metrics, consistent with known phenotypic limitations: w1118 males displayed restricted courtship due to visual impairment (Krstic et al., 2013), while y1 males demonstrated suboptimal mating performance, aligning with prior reports of courtship deficits (Drapeau et al., 2006).

Behavioral assays of Drosophila males under different rearing conditions.

(A) CL of Canton-S, w1118, y1, and Oregon-R males. (B) CI of Canton-S, w1118, y1, and Oregon-R males. (C) MD of Canton-S, w1118, y1, and Oregon-R males. For detailed methods, see the background for a detailed description of the mating assays used in this study. (D) CL of group-housed and single-reared Canton-S males. (E) CI of group-housed and single-reared Canton-S males. (F) LMD assays of Canton-S males. In the MD assays, white data points denote males that were group-reared (or sexually naïve), whereas blue data points signify males that were singly reared. The dot plots represent the MD of each male fly. The mean value and standard error are labeled within the dot plot. (G) CL of naive and sexually experienced Canton-S males. (H) CI of naive and sexually experienced Canton-S males. (I) SMD assays of Canton-S males. White data points represent sexually-naïve males and pink data points represent sexually-experienced ones.

The software further validated established behavioral paradigms—Longer Mating Duration (LMD) (Kim et al., 2013b, 2012b) and Shorter Mating Duration (SMD) (Lee et al., 2023b)—using Canton-S males subjected to distinct pre-experimental conditioning. DrosoMating accurately detected expected differences in CL, CI, and MD between LMD (group vs. single housing) and SMD (sexually naive vs. sexually experienced) conditions (Fig. 3D–I), confirming its precision in resolving complex, paradigm-specific mating dynamics.

Quantification of Learning-Memory Dynamics and Neural Circuit Integration Using DrosoMating

DrosoMating facilitates the investigation of learning and memory consolidation in Drosophila through quantifiable behavioral metrics. The Learning Index (LI)—calculated as (Fig. 4A), where CI denotes Courtship Index—provides a standardized measure of experience-dependent behavioral modulation. This transformation of abstract behavior into discrete parameters enables rigorous analysis of neural and behavioral plasticity.

behavioral assays of learning in Drosophila which combined with temperature-dependent temporal activation/inhibition methods.

(A) Learning Index (LI) calculation and experimental design for temperature-dependent activation/inhibition of neural circuits in adult Drosophila. This figure illustrates the experimental paradigm for calculating the LI using CI measurements from sham and trained group. The protocol involves rearing flies at 20°C for four days followed by a one-day exposure to 29°C to activate/inhibit temperature-dependent, genetically-encoded neuronal modulators (for example, shits as inhibitor or TrpA1 as activator). (B) CI of training and sham training MB-247-GAL4/UAS-GFP, MB-247-GAL4/UAS-TrpA1 and MB-247-GAL4/UAS-shits flies. (C) CI of different training days (D1: 1 day, D2: 2 days, D3: 3 days) and sham trained Canton-S flies. (D) CI of different training days (D1: 1day, D2: 2 days, D3: 3days) and sham training orbΔ/orbΔQ flies. (E) LI of Canton-S and orbΔ/orbΔQ males. (F) CI of training and sham training rut2080; UAS-rut, rut2080; GAL4ok107, rut2080; GAL4ok107/UAS-rut flies. (G) LI of rut2080; UAS-rut, rut2080; GAL4ok107, rut2080; GAL4ok107/UAS-rut flies.

The software integrates seamlessly with neural manipulation techniques. For acute temporal control, thermogenetic tools (UAS-shibirets(Kitamoto, 2001), UAS-TrpA1 (Kang et al., 2012)) permit precise activation or silencing of circuits during courtship (Fig. 4B and S1A). Targeted perturbation of mushroom body Kenyon cells, critical for memory processing, revealed immediate behavioral consequences (Fig. S1B–C), elucidating their role in behavioral adaptation.

Established courtship conditioning paradigms (Gil-Martí et al., 2023b; Griffith and Ejima, 2009a; Kamyshev et al., 1999; Reza et al., 2013; Wolf et al., 1998) are robustly replicated, validating DrosoMating’s fidelity in experience-driven behavior. (Fig. 4C). Mutant analyses revealed orb2 (Fig. 4D–E) and rut (Fig. 4F-G) dependence of experience-induced CI reduction, consistent with known memory pathways.”

Chronic neural interventions are enabled via tub-GAL80ts-mediated adult-specific silencing or knockdown (Fig. S1D–F). This approach reveals enduring impacts of sustained circuit modulation on courtship plasticity and memory consolidation. Collectively, DrosoMating bridges behavioral quantification and mechanistic inquiry, advancing exploration of learning-memory interplay through precise, scalable experimental frameworks.

Methods and Materials

Fly Stocks and Husbandry

Drosophila stocks were maintained on standard cornmeal-agar food at 25°C and 60-70% relative humidity, with a 12-hour light/dark cycle. When suppression of temperature-sensitive transgene activity was required (e.g., for UAS-shibirets or UAS-TrpA1 experiments), flies were reared at 19-20°C throughout development. Because developmental duration at 19-20°C is approximately twice that at 25°C, experimental scheduling was adjusted to accommodate the delayed eclosion. For the activation of transgenes was achieved by transferring flies to 29°C (shibirets) or 28°C (TrpA1) for 30 mins prior to experiments. To eliminate confounding variables in behavioral assays, control and experimental groups were always raised under identical environmental conditions, including temperature, humidity, and light cycle.

Environmental Control System

The controlled-environment chamber is a dedicated fly room equipped to maintain standardized experimental conditions. It features a precision temperature control unit, which is a programmable apparatus for maintaining constant temperature with an accuracy of ±0.5°C. The chamber also includes automated humidity regulation through a humidifier or climate control system to stabilize relative humidity, typically within the range of 60–70%. Additionally, a programmable LED lighting system with full-spectrum arrays and an integrated timer is used to simulate circadian cycles, allowing for adjustable day/night duration and light intensity.

Mating Behavior Measurement System Setup

Behavioral observation arena: Customizable fly chamber, constructed with [material, e.g., acrylic/polycarbonate] and dimensions [1 × 1 × 0.2 cm] to standardize fly interactions (Fig. 1D-E).

Sex-separation films: Non-permeable acrylic partitions (0.5 mm thickness) to isolate male and female cohorts prior to trials. Customized apparatus components can be fabricated by precision cutting of OHP (overhead projector) film sheets.

Uniform illumination system: Programmable LED panel array (full-spectrum, 6500K color temperature) positioned to optimize contrast for automated tracking software.

High-resolution imaging: OPPO A72 5G main camera (4K resolution, 30 fps) mounted on an adjustable stabilizer for consistent video capture.

Mating Behavior Measurement Preparations

This protocol outlines the collection of male and virgin female Drosophila melanogaster and details the standardized mating behavior assay. To ensure experimental reproducibility, each cohort (genotype or condition) must include ≥100 individuals per sex, with final mating counts reaching at least 20 successful pairs (ideally 70-80) within a 2-hour observation window. The protocol spans developmental stages from egg to adult emergence, with adults requiring 5 days post-eclosion to attain sexual maturity. Flies are reared under controlled environmental conditions (25°C, 60-70% relative humidity, 12-hour light/dark cycle) to minimize physiological variability. To avoid confounding effects of anesthesia on mating behavior, CO2 exposure was strictly minimized during fly handling and sorting; brief, low-intensity exposure was reserved for initial virgin collection and before starting behavioral assays.

Preparation of experimental male flies (15–20 days)

Raise flies of the desired genotype (wild-type, mutants, or crosses such as GAL4/UAS combinations) in bottles at the specified temperature (standard: 25°C; alternative: 19°C or 29°C as required) under a 12:12 h light:dark cycle. Maintain 2–3 bottles per genotype, each containing 25 females and 15 males, to yield 25–30 flies per day. Transfer flies to fresh food every 2–3 days to prevent overcrowding.

Post-eclosion handling

Collect newly eclosed males (0–8 h post-eclosion) under brief CO₂ anesthesia. Isolate individual males in vials and age them for 5 days at 25°C (or 9 days at 18°C) to ensure complete brain maturation. Clear existing adults from culture vials daily using CO₂ to ensure synchronized collection of newly eclosed males. Group males into cohorts of 40 and rear in vials for 5–9 days to optimize sexual maturity while preserving mating efficiency.

Large-scale virgin female collection for high-throughput genetic screening

The Df(1)Exel6234/Yhs-hid (Bloomington Stock Center #7708 with #8846) or the SPRattP/Yhs-hid strain (Bloomington Stock Center #84576 with #8846) were used in this study. Both strains carry a null mutation in the sex-peptide receptor (SPR) gene (Yapici et al., 2008), which disrupts post-mating female refractoriness, thereby enhancing receptivity to remating and heat-inducible hid expression eliminating male progeny and balancer heterozygotes. Maintain cultures below 20°C prior to egg collection to prevent premature hid-induced lethality in Yhs-hid containing stocks.Higher temperatures (>20°C) may activate leaky hs-hid expression, reducing male lethality efficiency during subsequent heat shock.

Place adult Df (1)Exel6234/Yhs-hid flies in fresh bottles for 24 h to deposit eggs at 22°C. Heat shock incubate eggs at 37°C for 75 min to trigger hid-mediated apoptosis of males and balancer heterozygotes. We recommend using a water bath heated to 37°C for efficient heat shock instead of simply placing the samples in a 37°C incubator. Collect surviving female progeny and rear in groups of 40 for 5–8 days under standard conditions at 25°C to produce age-matched, virgin females. Raising at 29°C can speed up the process of obtaining virgin females. Following heat shock treatment, inspect all eclosed adults to confirm complete absence of male progeny. The Df(1)Exel6234/Yhs-hidsystem should yield 100% female populations when properly executed.

Chamber specifications

Structure: Reinforced with four stainless steel columns (3 mm diameter) to ensure stability (Fig. 1D-E and 5A).

Material: Four transparent acrylic plates (180 × 75 mm) with alternating thicknesses (1 cm and 0.3 cm), assembled in a “thick-thin-thin-thick” sequence (Fig. 1F and 5C).

Layout: 36 individual chambers arranged in 3×4 clusters (12 chambers per cluster).

Chamber spacing: 1 mm between individual chambers; 2 cm between clusters. Positioned 3 cm from the chamber’s long edge and 1 cm from the short edge (Fig. 1E-F).

Chamber dimensions are not fixed. By simply adjusting the magnification of the recording lens, users can re-scale the field of view and then refine the x (horizontal) and y (vertical) grid values in the DrosoMating interface to enclose the entire fly activity area within the corresponding wells. After this one-step calibration, the pipeline quantifies mating metrics (CL, CI, MD, etc.) with identical accuracy across different chamber sizes.

Setup of mating assay chamber apparatus and architecture of workstation.

(A) Disassembly of the chamber apparatus. Each chamber comprises a roof, two activity areas, and a floor integrated with four pillars. (B) Workstation for the assemble mating assay chambers. Essential required equipments for chamber assembly are shown: a stereomicroscope with an integrated light source, carbon dioxide anesthesia apparatus, a soft brush with feather on the opposite end, chamber apparatus, and fly containing vials. (C) 3D schematic of the mating chamber assembly. Assembly from bottom to top as shown. Briefly, first assemble the floor with one activity area 2 for placing females, then cover the females and separate activity area 1 from activity area 2 using the sex-separation slider. Next, position males and secure the chambers with paper tape. (D) The sex-separation slider is positioned between activity area 1 and activity area 2. (E) The mating chamber assembled with paper tape; the sex-separation slider should be marked with a pen to prevent forgetting to remove it. (F) 3D schematic of single unit of chamber, can hold up to 36 pairs of fruit flies simultaneously. (G) Multiple chambers are set up with four chambers vertically aligned on an LED backlight, allowing the accommodation of 144 pairs of fruit flies across all chambers. (H) Video recording setup. Position a 4K/1080p resolution camera or a smartphone with matching video capabilities above the chambers, turn on the LED backlight, and remove sex-separation slider before recording. (I) Setup for recording mating behavior using a camera mounted on a stand, positioned vertically above chambers arranged on a size-matched LED backlight to ensure uniform illumination. The camera is aligned to capture the entire chamber array, enabling simultaneous recording of mating interactions under controlled lighting conditions.

Software implementation

Automated analysis is performed using the DrosoMating pipeline (GitHub https://github.com/hcls-kimlab/DrosoMating), the raw code is under win-ff module. which extracts courtship and mating metrics from video recordings. (Linux version: https://github.com/liuxiao916/Fruit_Flies)

Mating Behavior Assay Step-by-step Procedures

Fly preparation

On day 5 post-eclosion, select healthy males (of the desired genotype) and virgin females. Lightly anesthetize flies using ice or CO₂ pad (avoid prolonged exposure to prevent stress) (Fig. 5B). To minimize behavioral perturbations, flies should only be anesthetized at this step. For brief immobilization (e.g., sorting or transferring individuals), use CO₂ anesthesia (≤30 sec exposure). To avoid bias, all experiments should be done blindly. Label fly tubes with random numbers instead of genotypes and keep a separate list matching numbers to genotypes. Only check which numbers correspond to which genotypes after finishing all data analysis.

Chamber assembly and fly loading

Using a soft brush, gently place a single female into a chamber hole in Floor (Activity Area 2) (Fig. 5B). Cover the chamber with a transparent film to prevent physical contact while allowing visual/chemical interactions. Align Activity Area 1 on top of the film (Fig. 5C). Introduce a single male into the opposite side of the film using the transparent film as a slider to minimize CO₂ exposure (Fig. 5D). Limit male exposure to CO₂ during this step to avoid behavioral artifacts. Seal the chamber securely with Roof using disposable paper tape (Fig. 5E).

Recovery and Mating Initiation

Transfer the assembled chamber to a 25°C incubator (humidity controlled at 60–70% RH) for 90 min to ensure flies recover fully from handling stress. After recovery, gently remove the transparent film to initiate physical mating (Fig. 5E and F).

Video Recording Setup

Position an adjustable LED panel beneath the chamber to ensure uniform illumination (Fig. 5G). Record mating behavior at 4K (1080p also meets the requirements) resolution for 1.5–2 hours (adjust duration as needed) (Fig. 5H). A minimal setup for capturing fruit fly mating behaviors can be rapidly established without extensive external facility requirements. The essential components include a camera stand, an LED light panel, and a camera capable of recording at a resolution and frame rate of 4K (60 fps). (Fig. 5I).

Analysis of Drosophila Mating Video Data Analysis and Statistics

The use of the Windows version of Drosomating

Prepare computer equipped either Windows (or Linux) PC with DrosoMating (ff-chose.exe) installed. The original code instructions for use are provided under the win-ff module.

Transfer video files into the computer to be analyzed. Launch DrosoMating and open the video file.

Enter the number of boards and velocity parameter. After entering the information, click “Enter” to proceed to the next step. (Fig. 1A)

. Select the four steel marker pillars in clockwise order (Fig. 1A). Adjust the selection sequence based on the video orientation. For each plate, a preview image will appear. Click on three distinct flies to enhance tracking accuracy (Fig. 1B).

After selecting all regions, return to the DrosoMating home interface. Adjust x (horizontal) and y (vertical) values to align grids with chamber holes. Fine-tune fly detection by modifying the s value (sensitivity threshold) until all flies are correctly identified. Use the grid to segment individual flies (as shown in Fig. 1C).

Run the analysis (Save the processed video but not recommended for routine use).

Export raw data as a CSV or Excel file. Open it in spreadsheet software for further processing.

The usage instructions for the Linux version of Drosomating are presented in Fig. S2

Statistical Analysis

To ensure robust statistical analysis, each experimental group included at least 100 male flies (naïve, sexually experienced, or singly reared). Internal controls were incorporated into every experiment as recommended by Bretman et al. (2011) (Bretman et al., 2011). Normality of the mating duration data was confirmed using the Kolmogorov-Smirnov test (p>0.05). For group comparisons, two-sided Student’s t-tests were applied to calculate significance levels (****p<0.0001, ***p<0.001, **p<0.01, * p<0.05), while estimation statistics (Claridge-Chang and Assam, 2016) were used to visualize effect sizes, mean differences, and precision, avoiding reliance solely on null hypothesis testing. All analyses, including data plotting, were performed using GraphPad Prism software.

Resource availability

Lead contact

Further information and requests for resources and reagents should be directed to and will be fulfilled by the lead contact, Woo Jae Kim (wkim@hit.edu.cn)

Materials availability

This study did not generate new, unique reagents.

Discussion

Advancing Courtship Behavior Analysis in Drosophila: From Manual Tracking to computer-vision classifier

The study of male Drosophila mating behavior represents a cornerstone in neurogenetics, providing critical insights into the neural and molecular mechanisms underlying innate social behaviors (Dauwalder, 2008; Dukas, 2020). Historically, the CI—quantifying the proportion of time spent in courtship activities—has served as a fundamental metric for evaluating mating activity (Greenspan and Ferveur, 2000a; Griffith and Ejima, 2009b; Pavlou and Goodwin, 2013; Sokolowski, 2001; Yamamoto and Koganezawa, 2013). However, traditional methodologies relied on labor-intensive manual observation, limiting their capacity to capture dynamic behavioral repertoires such as orientation, wing extension (“courtship song”), and copulation sequences (Dauwalder, 2008; Dukas, 2020). While machine vision systems like Dankert et al.’s automated platform advanced quantification of social interactions, their primary application skewed toward male-male aggression rather than nuanced courtship dynamics (Dankert et al., 2009b). Concurrently, research into genetic and environmental modulators—including fruitless (fru) and doublesex (dsx) gene pathways, cuticular hydrocarbons, and sensory cues—highlighted the need for scalable, high-resolution tools to dissect multimodal influences on behavior.

Technical Innovation and Validation

DrosoMating addresses these gaps through an integrated framework combining modular environmental chambers, automated video tracking, and machine-based annotation. This system minimizes manual intervention while capturing temporally precise metrics such as CL, CI, and MD. Validation against ground-truth manual scoring demonstrated exceptional accuracy (98–99% agreement; error rates <0.05%), underscoring reliability for high-throughput screens. Crucially, the platform detected subtle behavioral variations across genotypes—e.g., e.g., diminished courtship vigor in w1118 and y1 mutants—and environmental manipulations such as group-rearing or prior sexual experience, which respectively shortened mating duration by ∼8–12 %. This sensitivity enables investigations into adaptive trade-offs, learning phenotypes, and circuit-level mechanisms. For instance, DrosoMating replicated classic deficits in aversive learning following mushroom body ablation and courtship impairments in rutabaga learning mutants, corroborating its utility for behavioral neuroscience (Levin et al., 1992).

Limitations and Future Directions

Despite its advantages, DrosoMating currently lacks granularity for posture-specific interactions (e.g., “wing threat” during aggression or “circling” during courtship) and multisensory integration. Future iterations could incorporate 3D pose estimation algorithms (e.g., DeepLabCut) or integrate JAABA-based classifiers to expand action repertoires (Kabra et al., 2013). Environmental control—though essential for standardization—may limit ecological validity;

eLife Assessment

This useful paper describes a software tool, "DrosoMating", which allows automated, high-throughput quantification of 6 common metrics of courtship and mating behaviors in Drosophila melanogaster. The validity of the tool is quite convincingly demonstrated by comparing expert human assessments with those made by DrosoMating. The work, however, does not address how DrosoMating compares with or advances on other existing tools for exactly the same purpose, whether it can be used for studies of other Drosophila species, and/or whether finer aspects of courtship response timing - which depend on proximal female signals to the male - could be extracted with more detailed analyses. Some additional statistical analyses would also help further strengthen the authors' current conclusions.

Reviewer #1 (Public review):

Summary:

The study of Drosophila mating behaviors has offered a powerful entry point for understanding how complex innate behaviors are instantiated in the brain. The effectiveness of this behavioral model stems from how readily quantifiable many components of the courtship ritual are, facilitating the fine-scale correlations between the behaviors and the circuits that underpin their implementation. Detailed quantification, however, can be both time-consuming and error-prone, particularly when scored manually. Song et al. have sought to address this challenge by developing DrosoMating, software that facilitates the automated and high-throughput quantification of 6 common metrics of courtship and mating behaviors. Compared to a human observer, DrosoMating matches courtship scoring with high fidelity. Further, the authors demonstrate that the software effectively detects previously described variations in courtship resulting from genetic background or social conditioning. Finally, they validate its utility in assaying the consequences of neural manipulations by silencing Kenyon cells involved in memory formation in the context of courtship conditioning.

Strengths:

(1) The authors demonstrate that for three key courtship/mating metrics, DrosoMating performs virtually indistinguishably from a human observer, with differences consistently within 10 seconds and no statistically significant differences detected. This demonstrates the software's usefulness as a tool for reducing bias and scoring time for analyses involving these metrics.

(2) The authors validate the tool across multiple genetic backgrounds and experimental manipulations to confirm its ability to detect known influences on male mating behavior.

(3) The authors present a simple, modular chamber design that is integrated with DrosoMating and allows for high-throughput experimentation, capable of simultaneously analyzing up to 144 fly pairs across all chambers.

Weaknesses:

(1) DrosoMating appears to be an effective tool for the high-throughput quantification of key courtship and mating metrics, but a number of similar tools for automated analysis already exist. FlyTracker (CalTech), for instance, is a widely used software that offers a similar machine vision approach to quantifying a variety of courtship metrics. It would be valuable to understand how DrosoMating compares to such approaches and what specific advantages it might offer in terms of accuracy, ease of use, and sensitivity to experimental conditions.

(2) The courtship behaviors of Drosophila males represent a series of complex behaviors that unfold dynamically in response to female signals (Coen et al., 2014; Ning et al., 2022; Roemschied et al., 2023). While metrics like courtship latency, courtship index, and copulation duration are useful summary statistics, they compress the complexity of actions that occur throughout the mating ritual. The manuscript would be strengthened by a discussion of the potential for DrosoMating to capture more of the moment-to-moment behaviors that constitute courtship. Even without modifying the software, it would be useful to see how the data can be used in combination with machine learning classifiers like JAABA to better segment the behavioral composition of courtship and mating across genotypes and experimental manipulations. Such integration could substantially expand the utility of this tool for the broader Drosophila neuroscience community.

(3) While testing the software's capacity to function across strains is useful, it does not address the "universality" of this method. Cross-species studies of mating behavior diversity are becoming increasingly common, and it would be beneficial to know if this tool can maintain its accuracy in Drosophila species with a greater range of morphological and behavioral variation. Demonstrating the software's performance across species would strengthen claims about its broader applicability.

Reviewer #2 (Public review):

This paper introduces "DrosoMating," an integrated hardware and software solution for automating the analysis of male Drosophila courtship. The authors aim to provide a low-cost, accessible alternative to expensive ethological rigs by utilizing a custom acrylic chamber and smartphone-based recording. The system focuses on quantifying key temporal metrics-Courtship Index (CI), Copulation Latency (CL), and Mating Duration (MD)-and is applied to behavioral paradigms involving memory mutants (orb2, rut).

The development of open-source behavioral tools is a significant contribution to neuroethology, and the authors successfully demonstrate a system that simplifies the setup for large-scale screens. A major strength of the work is the specific focus on automating Copulation Latency and Mating Duration, metrics that are often labor-intensive to score manually.

However, there are several limitations in the current analysis and validation that affect the strength of the conclusions:

First, the statistical rigor requires substantial improvement. The analysis of multi-group experiments (e.g., comparing four distinct strains or factorial designs with genotype and training) currently relies on multiple independent Student's t-tests. This approach is statistically invalid for these experimental designs as it inflates the family-wise Type I error rate. To support the claims of strain-specific differences or learning deficits, the data must be analyzed using Analysis of Variance (ANOVA) to properly account for multiple comparisons and to explicitly test for interaction effects between genotype and training conditions.

Second, the biological validation using w1118 and y1 mutants entails a potential confound. The authors attribute the low Courtship Index in these strains to courtship-specific deficits. However, both strains are known to exhibit general locomotor sluggishness (due to visual or pigmentation/behavioral defects). Since "following" behavior is likely a component of the Courtship Index, a reduction in this metric could reflect a general motor deficit rather than a specific lack of reproductive motivation. Without controlling for general locomotion, the interpretation of these behavioral phenotypes remains ambiguous.

Third, the benchmarking of the system is currently limited to comparisons against manual scoring. Given that the field has largely adopted sophisticated open-source tracking tools (e.g., Ctrax, FlyTracker, JAABA), the utility of DrosoMating would be better contextualized by comparing its performance - in terms of accuracy, speed, or identity maintenance - against these existing automated standards, rather than solely against human observation.

Finally, the visual presentation of the data hinders the assessment of the system's temporal precision. While the system is designed to capture time-resolved metrics, the results are presented primarily as aggregate bar plots. The absence of behavioral ethograms or raster plots makes it difficult to verify the software's ability to accurately detect specific transitions, such as the exact onset of copulation.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The study of Drosophila mating behaviors has offered a powerful entry point for understanding how complex innate behaviors are instantiated in the brain. The effectiveness of this behavioral model stems from how readily quantifiable many components of the courtship ritual are, facilitating the fine-scale correlations between the behaviors and the circuits that underpin their implementation. Detailed quantification, however, can be both time-consuming and error-prone, particularly when scored manually. Song et al. have sought to address this challenge by developing DrosoMating, software that facilitates the automated and high-throughput quantification of 6 common metrics of courtship and mating behaviors. Compared to a human observer, DrosoMating matches courtship scoring with high fidelity. Further, the authors demonstrate that the software effectively detects previously described variations in courtship resulting from genetic background or social conditioning. Finally, they validate its utility in assaying the consequences of neural manipulations by silencing Kenyon cells involved in memory formation in the context of courtship conditioning.

Strengths:

(1) The authors demonstrate that for three key courtship/mating metrics, DrosoMating performs virtually indistinguishably from a human observer, with differences consistently within 10 seconds and no statistically significant differences detected. This demonstrates the software's usefulness as a tool for reducing bias and scoring time for analyses involving these metrics.

(2) The authors validate the tool across multiple genetic backgrounds and experimental manipulations to confirm its ability to detect known influences on male mating behavior.

(3) The authors present a simple, modular chamber design that is integrated with DrosoMating and allows for high-throughput experimentation, capable of simultaneously analyzing up to 144 fly pairs across all chambers.

Weaknesses:

(1) DrosoMating appears to be an effective tool for the high-throughput quantification of key courtship and mating metrics, but a number of similar tools for automated analysis already exist. FlyTracker (CalTech), for instance, is a widely used software that offers a similar machine vision approach to quantifying a variety of courtship metrics. It would be valuable to understand how DrosoMating compares to such approaches and what specific advantages it might offer in terms of accuracy, ease of use, and sensitivity to experimental conditions.

(2) The courtship behaviors of Drosophila males represent a series of complex behaviors that unfold dynamically in response to female signals (Coen et al., 2014; Ning et al., 2022; Roemschied et al., 2023). While metrics like courtship latency, courtship index, and copulation duration are useful summary statistics, they compress the complexity of actions that occur throughout the mating ritual. The manuscript would be strengthened by a discussion of the potential for DrosoMating to capture more of the moment-to-moment behaviors that constitute courtship. Even without modifying the software, it would be useful to see how the data can be used in combination with machine learning classifiers like JAABA to better segment the behavioral composition of courtship and mating across genotypes and experimental manipulations. Such integration could substantially expand the utility of this tool for the broader Drosophila neuroscience community.

(3) While testing the software's capacity to function across strains is useful, it does not address the "universality" of this method. Cross-species studies of mating behavior diversity are becoming increasingly common, and it would be beneficial to know if this tool can maintain its accuracy in Drosophila species with a greater range of morphological and behavioral variation. Demonstrating the software's performance across species would strengthen claims about its broader applicability.

Reviewer #2 (Public review):

This paper introduces "DrosoMating," an integrated hardware and software solution for automating the analysis of male Drosophila courtship. The authors aim to provide a low-cost, accessible alternative to expensive ethological rigs by utilizing a custom acrylic chamber and smartphone-based recording. The system focuses on quantifying key temporal metrics-Courtship Index (CI), Copulation Latency (CL), and Mating Duration (MD)-and is applied to behavioral paradigms involving memory mutants (orb2, rut).

The development of open-source behavioral tools is a significant contribution to neuroethology, and the authors successfully demonstrate a system that simplifies the setup for large-scale screens. A major strength of the work is the specific focus on automating Copulation Latency and Mating Duration, metrics that are often labor-intensive to score manually.

However, there are several limitations in the current analysis and validation that affect the strength of the conclusions:

First, the statistical rigor requires substantial improvement. The analysis of multi-group experiments (e.g., comparing four distinct strains or factorial designs with genotype and training) currently relies on multiple independent Student's t-tests. This approach is statistically invalid for these experimental designs as it inflates the family-wise Type I error rate. To support the claims of strain-specific differences or learning deficits, the data must be analyzed using Analysis of Variance (ANOVA) to properly account for multiple comparisons and to explicitly test for interaction effects between genotype and training conditions.

Second, the biological validation using $w^{1118}$ and $y^1$ mutants entails a potential confound. The authors attribute the low Courtship Index in these strains to courtship-specific deficits. However, both strains are known to exhibit general locomotor sluggishness (due to visual or pigmentation/behavioral defects). Since "following" behavior is likely a component of the Courtship Index, a reduction in this metric could reflect a general motor deficit rather than a specific lack of reproductive motivation. Without controlling for general locomotion, the interpretation of these behavioral phenotypes remains ambiguous.

Third, the benchmarking of the system is currently limited to comparisons against manual scoring. Given that the field has largely adopted sophisticated open-source tracking tools (e.g., Ctrax, FlyTracker, JAABA), the utility of DrosoMating would be better contextualized by comparing its performance - in terms of accuracy, speed, or identity maintenance - against these existing automated standards, rather than solely against human observation.

Finally, the visual presentation of the data hinders the assessment of the system's temporal precision. While the system is designed to capture time-resolved metrics, the results are presented primarily as aggregate bar plots. The absence of behavioral ethograms or raster plots makes it difficult to verify the software's ability to accurately detect specific transitions, such as the exact onset of copulation.

We sincerely thank the reviewers for their constructive and detailed feedback, which has substantially improved the clarity and rigor of our work. Below is a summary of the major revisions.

(1) Comparison with existing tools. We added a new main figure (Figure 5) and Table 1 systematically benchmarking DrosoMating against Ctrax and FlyTracker on identical low-quality, high-throughput mating videos. Both conventional tools failed under our recording conditions: Ctrax showed severe segmentation instability and fragmented trajectories, while FlyTracker frequently crashed during feature computation. These results demonstrate that DrosoMating's state-detection approach bypasses the pose-tracking limitations that impair established pipelines. We also revised the Discussion to clarify that DrosoMating is specialized for robust extraction of mating timing metrics from low-quality videos, not a replacement for general-purpose tracking tools.

(2) Statistical analysis. Following Reviewer #2's recommendation, we re-analyzed Figure 3 using one-way ANOVA with Tukey's HSD for strain comparisons, and Figure 4 using two-way ANOVA with Sidak's post-hoc test for learning assays, including the critical Genotype by Training interaction term.

(3) New supplementary data. We added Figure S4 showing basal locomotor velocity of single-housed males across all four strains to decouple motor defects from courtship deficits in w1118 and y1 mutants. We also added Figure S3 with behavioral ethograms for individual flies to visually demonstrate the system's temporal resolution and detection accuracy.

(4) Additional improvements. We standardized statistical reporting across all figure legends, explicitly defined the segmentation threshold parameter s in the Methods, consistently used "Mating Duration (MD)" throughout, standardized MB247-GAL4 labeling, clarified that occluded frames are retained for mating-state detection via merged-contour analysis, and corrected minor errors including duplicate references and unnecessary quotation marks.

We believe these revisions substantially strengthen the manuscript and clearly position DrosoMating within the existing ecosystem of behavioral analysis tools.

We remain grateful for the valuable feedback from both reviewers and the editorial team, and we hope the revised version meets the standards for publication in eLife.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

(1) It's difficult to assess the utility of this tool in relation to the variety of alternative methods available for the automated scoring of courtship behavior, some of which offer more granular behavioral data than what DrosoMating has been presented to produce. A direct comparison to other approaches (e.g., FlyTracker, DeepLabCut, SLEAP) would be highly valuable for understanding what use cases DrosoMating is best suited for. Ideally, this analysis would include (i) a comparison of the accuracy of scoring courtship metrics and constituent behaviors, (ii) an assessment of the sensitivity of the various software to different experimental conditions, such as lighting and alternative chamber designs, and (iii) testing across Drosophila species, which could help highlight the particular strengths of DrosoMating over more established methods. At a minimum, a table comparing key features, requirements, and capabilities would help position DrosoMating within the existing ecosystem of tools.

We sincerely appreciate the reviewer’s insightful comments on benchmarking DrosoMating against existing tools for automated courtship behavior analysis. We fully agree that direct comparisons are essential to clarify the unique advantages and ideal use cases of DrosoMating. We have extensively revised the manuscript by adding comparative experiments, a new figure（Figure 5) and comparative table (Table1), and expanded descriptions, as detailed below.

(1) Direct comparison with conventional tracking tools

We systematically tested Ctrax and FlyTracker on the same low-quality, high-throughput mating videos used for DrosoMating. The corresponding results are presented in Figure 5.

Ctrax showed severe segmentation instability, including over-segmentation, under-segmentation, and fragmented trajectories, and failed to stably detect two flies.

FlyTracker was able to generate background models and showed partially improved segmentation after threshold adjustment, but frequently crashed during feature computation and could not complete the end-to-end analysis pipeline.

These results confirm that tracking-based tools are vulnerable to low contrast, chamber artifacts, and prolonged male–female overlap, whereas DrosoMating bypasses pose tracking and directly detects mating states.

The revised manuscript text is as follows (line 265-342):

"DrosoMating is more compatible with low-quality mating videos than conventional tracking-based pipelines.

To evaluate whether conventional fly-tracking pipelines could be used as alternative tools for extracting mating-duration metrics, we tested Ctrax and FlyTracker on representative low-quality single-chamber mating videos recorded under our standard high-throughput conditions (Fig. 5). Ctrax is designed to estimate the position and orientation of multiple walking flies while maintaining individual identities over time (Branson et al., 2009) (https://ctrax.sourceforge.net/), whereas FlyTracker aims to track fly pose, including position, orientation, body size, wing and leg positions, and to generate trajectory- and feature-based outputs for downstream behavior analysis (Eyjolfsdottir et al., 2014). Because these programs were not readily compatible with our full high-throughput behavioral recording setup, we first cropped the original videos and tested single-chamber videos containing one male and one female fly.

Using Ctrax, we observed that target detection was highly variable even within single-chamber videos. In representative frames, Ctrax could occasionally identify the two flies correctly (Fig. 5A). However, imperfect segmentation was frequently observed. In some frames, parts of the fly body were detected as additional targets, resulting in over-segmentation and an apparent increase in the number of detected flies (Fig. 5B, C). Conversely, when the male and female were close to each other or physically overlapped, the two animals were sometimes detected as a single target (Fig. 5D). These examples indicate that Ctrax detection was sensitive to the low contrast and overlapping fly bodies present in our mating videos. We then adjusted the Ctrax detection threshold to improve segmentation quality. Although threshold optimization improved detection in some frames, abnormal detections remained evident across randomly sampled frames (Fig. 5E). For example, some frames still showed incorrect target numbers, and a severe segmentation failure was observed in the lower-right example, where the detected objects did not correspond to two clearly separable flies. Thus, even after parameter optimization, Ctrax did not consistently maintain the expected two-target detection state in single-chamber mating videos.

The instability of Ctrax detection was also reflected in the trajectory output. In the first 500 frames, the generated trajectories were fragmented into multiple colored track segments rather than two continuous trajectories corresponding to the male and female (Fig. 5F). In addition, some trajectories extended outside the chamber boundary, indicating tracking errors and identity instability. Consistently, frame-by-frame quantification of detected target number showed that the detected object count did not remain stable at the expected value of two flies per chamber (Fig. 5G). Because Ctrax failed to maintain stable two-fly detection and continuous trajectories under these video conditions, its output could not be reliably used for downstream extraction of copulation latency or mating duration.

We next tested FlyTracker on the same type of cropped single-chamber mating videos. FlyTracker was developed to track multiple flies by estimating body position, orientation, size, wing and leg positions, and by maintaining fly identities across video frames; it also outputs per-frame features such as velocity, facing angle, and wing-angle-related measurements for downstream behavioral analysis (Eyjolfsdottir et al., 2014). In our videos, FlyTracker was able to generate a background model, indicating that the program could recognize the overall imaging field and chamber background (Fig. 5H). However, during calibration and segmentation, the default threshold setting produced inconsistent detection results (Fig. 5I). Although some frames were segmented relatively well under the default threshold (Fig. 5J), these successful examples were not representative of the overall tracking process, and the program frequently terminated with runtime errors during tracking or downstream feature computation. To improve detection stability, we lowered the segmentation threshold. Under this adjusted setting, FlyTracker produced more complete fly masks in representative frames (Fig. 5K), and the diagnostic output showed that two flies were detected in many sampled frames (Fig. 5L). Nevertheless, detection remained unstable in some sampled frames, including frames in which zero or one fly was detected despite the expected two flies per chamber (Fig. 5L). The full pipeline ultimately failed during feature computation, producing a runtime error before complete tracking and feature outputs could be generated (Fig. 5M). Therefore, even after threshold adjustment, FlyTracker could not provide a stable end-to-end workflow for extracting mating-duration metrics from these low-quality mating videos.

This failure mode is relevant because FlyTracker depends on stable segmentation, identity maintenance, and per-frame feature extraction. In our assay videos, low contrast, chamber-edge artifacts, and prolonged male–female overlap during copulation interfered with these requirements. As a result, FlyTracker could occasionally identify the flies in individual frames, but it did not reliably complete the full analysis pipeline required for downstream behavioral quantification. This limitation is especially important for workflows such as JAABA, which use manually labeled examples to train behavior classifiers but still depend on upstream tracking-derived features. Thus, for our low-quality high-throughput mating recordings, FlyTracker-based analysis was substantially less practical than DrosoMating, which directly outputs mating-related timing metrics without requiring continuous high-fidelity two-fly pose tracking."

(2) Evaluation of accuracy, robustness, and cross-species potential

We addressed the three key points requested by the reviewer:

(i) Accuracy: DrosoMating achieved 98–99% agreement with manual scoring. Under our experimental conditions, neither Ctrax nor FlyTracker completed an end-to-end workflow capable of reliably extracting copulation latency and mating duration. Ctrax produced fragmented trajectories with unstable target counts, and FlyTracker terminated with runtime errors during feature computation.

(ii) Robustness: DrosoMating is highly robust to low-contrast lighting and common behavioral chamber setups, whereas conventional tools require high-quality videos and fail under fly occlusion.

(iii) Cross-species testing: We did not perform cross-species validation in this revision. DrosoMating relies on mating state detection rather than species-specific morphology, suggesting potential transferability, although this remains to be experimentally validated. We have therefore revised the text to avoid claiming demonstrated cross-species performance and now describe this as a potential future application.

The revised manuscript text is as follows (line 572-576):

"Cross-species testing was not performed in this study. Although DrosoMating's state-detection approach is morphology-agnostic and therefore theoretically applicable across Drosophila species, we have revised the text to avoid claiming demonstrated cross-species performance. Formal validation across diverse species remains a promising future direction."

We added Table1 to compare key features across DrosoMating, Ctrax, and FlyTracker, including low-quality video compatibility, multi-chamber support, robustness to fly overlap, direct output of mating timing metrics, accuracy.

(3) Clarification of ideal use cases

We revised the Discussion to emphasize that DrosoMating is not intended to replace general-purpose pose or tracking tools (e.g., FlyTracker, Ctrax) that provide fine-grained behavioral data.

Instead, it offers a specialized, robust, and high-throughput workflow for scenarios requiring efficient extraction of mating timing metrics from low-quality videos, where conventional pipelines often fail.

These revisions greatly improve the clarity and positioning of DrosoMating. We thank the reviewer for this valuable suggestion.

The revised manuscript text is as follows (line 578-620):

“Comparative Evaluation with Existing Courtship Analysis Tools

A major advantage of DrosoMating is its compatibility with low-quality, high-throughput mating videos. Conventional tracking-based tools such as Ctrax and FlyTracker are powerful for trajectory- and pose-based behavioral analysis, but they generally require stable object segmentation, identity maintenance, and reliable feature extraction across frames. Ctrax was designed to estimate the positions and orientations of multiple walking flies while maintaining their identities, whereas FlyTracker aims to track detailed fly pose and generate per-frame behavioral features. Under our recording conditions, these requirements were difficult to satisfy because the videos were low contrast and the male and female frequently overlapped during copulation.

In our tests, both tools showed limited compatibility with these videos. Even after cropping to single-chamber videos and adjusting detection parameters, Ctrax produced unstable target numbers, fragmented trajectories, and tracking errors. FlyTracker could generate a background model and occasionally segment flies successfully, but detection remained unstable and the full pipeline failed during feature computation. These issues are particularly relevant for copulation latency and mating duration analysis: during copulation, the male and female remain physically coupled for a long period, which makes identity-based tracking difficult. For these timing metrics, it is more important to robustly detect the onset and offset of the mating state than to reconstruct detailed individual trajectories. DrosoMating was purpose-built to address these specific challenges. It operates reliably on lower-quality video streams, requires no complex pre-processing or manual ROI definition, and is optimized for high-throughput multi-chamber analysis. While it does not offer the same level of pose or kinematic detail as other tools, it provides a unique solution for laboratories seeking a simple, fast, and robust pipeline to quantify core reproductive timing metrics—copulation latency (CL), courtship index (CI), and mating duration (MD)—without the overhead of more complex systems (Table 1).

Although we directly tested only Ctrax and FlyTracker, this limitation may also affect workflows that depend on upstream tracking-derived features. For example, JAABA uses tracking-derived features to train behavior classifiers, and DANCE, a recent Drosophila aggression and courtship pipeline, uses JAABA-based classifiers and lists FlyTracker and JAABA as required software. Thus, our conclusion is not that these tools are generally unsuitable for Drosophila behavior analysis, but that DrosoMating provides a more practical workflow for low-quality, high-throughput videos focused specifically on mating timing.

While DrosoMating currently prioritizes mating timing metrics over discrete behavioral classification, its modular architecture provides a foundation for future integration with behavior classifiers such as JAABA. Laboratories requiring granular behavioral elements—such as wing extension or circling—would benefit from an extended pipeline that exports per-frame kinematic features for downstream classifier training. Validating this integration represents a promising future direction to broaden the tool’s utility beyond core reproductive timing assays.”

We have also added this clarification to the Table 1 legend to avoid overgeneralizing the limitations of Ctrax and FlyTracker (line 786-790):

"DrosoMating was designed to extract mating-related timing metrics from high-throughput videos without requiring continuous two-fly identity tracking. Ctrax and FlyTracker were tested on cropped single-chamber videos from the same recording setup. Their limitations described here refer specifically to these low-quality mating videos and should not be interpreted as general limitations of the tools."

(2) The coarse nature of the metrics captured by DrosoMating may limit the usefulness of this tool for many researchers. Consider integrating DrosoMating with one or more behavioral classifiers (e.g., JAABA) and validating its performance to increase its utility across a wider range of uses. Demonstrating the feasibility of this would substantially increase the tool's appeal to those who need access to more discrete behavioral elements.

We thank the reviewer for this constructive suggestion. We agree that DrosoMating is currently optimized for rapid, high-throughput quantification of core mating timing metrics (CL, CI, and MD) rather than discrete behavioral classification (e.g., wing extension, circling, or aggressive postures). This reflects a deliberate design trade-off: by prioritizing robust state detection over detailed pose tracking, DrosoMating achieves reliable performance on low-quality videos where conventional identity-based pipelines fail.

We appreciate the reviewer’s vision for expanding the tool’s utility. While DrosoMating does not presently generate the per-frame kinematic features (position, orientation, wing angles, etc.) required as input for JAABA classifiers, its modular architecture and underlying video-processing framework provide a foundation for future integration. In the revised Discussion, we have clarified that extending DrosoMating to export trajectory-derived features compatible with downstream classifiers such as JAABA represents a promising future direction—one that would broaden its applicability to laboratories requiring granular behavioral elements, without necessitating a complete overhaul of the existing pipeline.

We hope this clarification addresses the reviewer’s concern and accurately reflects both the current capabilities and future potential of the tool.

The revised manuscript text is as follows (line 614-620):

"While DrosoMating currently prioritizes mating timing metrics over discrete behavioral classification, its modular architecture provides a foundation for future integration with behavior classifiers such as JAABA. Laboratories requiring granular behavioral elements—such as wing extension or circling—would benefit from an extended pipeline that exports per-frame kinematic features for downstream classifier training. Validating this integration represents a promising future direction to broaden the tool’s utility beyond core reproductive timing assays."

(3) Please clarify how DrosoMating handles fly identification during mating? I would think mounted flies would be occluded, and if these frames are excluded from analysis, it would be expected to skew mating duration scores. It would be helpful if the authors could discuss these details.

Occluded frames are excluded from CI calculation because individual courtship actions cannot be reliably assigned during prolonged overlap. However, these frames are not discarded from mating-duration analysis. Instead, prolonged merged contours are used as evidence for copulation-state detection, and the MD timer continues until physical separation:

(1) When two flies are separate, the system detects two contours; when they overlap during mounting, it detects one merged contour. Our code processes both cases, so occluded frames are retained, not skipped.

(2) To distinguish a single fly from two overlapping flies, we analyze the shape of the merged contour. Two overlapping flies produce a characteristically different aspect ratio (elongated shape) compared to one fly. This geometric cue is fed into the state classifier to label the frame as "mating."

(3) Because these frames are classified as mating rather than excluded, the mating duration timer runs continuously through the occlusion period. Thus, mating duration is not artificially shortened.

The high agreement between DrosoMating and manual scoring for mating duration (within 10 s, no significant difference) confirms that this approach does not introduce bias.

We revised the manuscript as follow (line 191-193):

"Occluded frames are excluded from CI calculation but retained for mating-state detection via merged-contour analysis, ensuring continuous MD measurement through copulation."

(4) Please indicate the statistical tests used in Figures 3, 4, and S1. What methods were used to address multiple hypothesis testing?

We thank the reviewer for this important comment. For comparisons between two groups, two-sided Student’s t-tests were used. For comparisons among three or more groups, one-way ANOVA followed by post-hoc tests were applied. For two-factor experimental designs, two-way ANOVA was used. We have clearly stated these statistical tests in the Statistical Analysis section and have added this information to the figure legends for Figures 3, 4, and S1 in the revised manuscript.

The revised "Statistical Analysis" section is as follows (line 512-524):

"To ensure robust statistical analysis, each experimental group included at least 100 male flies (naïve, sexually experienced, or singly reared). Internal controls were incorporated into every experiment as recommended by Bretman et al. (2011) (Bretman et al., 2011). Normality of the mating duration data was confirmed using the Kolmogorov-Smirnov test (p>0.05). For group comparisons, two-sided Student’s t-tests were applied to calculate significance levels (****p<0.0001, ***p<0.001, **p<0.01, * p<0.05), Comparisons among three or more groups were performed using one-way ANOVA with Tukey’s HSD post-hoc tests. Two-factor experimental comparisons were performed using two-way ANOVA followed by Sidak’s post-hoc test. while estimation statistics (Claridge-Chang and Assam, 2016) were used to visualize effect sizes, mean differences, and precision, avoiding reliance solely on null hypothesis testing. All analyses, including data plotting, were performed using GraphPad Prism software."

(5) Line 43: "High resolution video tracking enables...".

We thank the reviewer for noting this imprecise expression. We have revised the statement to accurately describe that our method uses image-based video analysis to identify courtship and copulation events and quantify their temporal parameters. The description has been corrected in the revised manuscript line 44-46:

"Our image-based video analysis enables precise identification of courtship and copulation events, as well as quantification of their timing and duration under controlled conditions. "

(6) Line 51: remove quotation mark.

We thank the reviewer for the careful correction. The unnecessary quotation mark at Line 51 has been removed in the revised manuscript.

(7) Line 107: Chen et al, 2024 is listed twice in the references.

We thank the reviewer for the careful correction. The duplicate reference of Chen et al., 2024 has been removed from the reference list in the revised manuscript.

(8) Line 242: remove quotation mark.

We thank the reviewer for the careful correction. The unnecessary quotation mark at Line 242 has been removed in the revised manuscript.

Reviewer #2 (Recommendations for the authors):

(1) Please re-analyze the data in Figures 3 and 4 using ANOVA followed by appropriate post-hoc tests (e.g., Tukey's HSD). Specifically, use a One-way ANOVA for strain comparisons in Figure 3 and a Two-way ANOVA for the learning assays in Figure 4. The interaction term (Genotype $\times$ Training) is critical for demonstrating specific learning deficits. Update the "Statistical Analysis" section and figure legends accordingly.

We appreciate the reviewer’s recommendation. We have re-analyzed the data in Figures 3 and 4 using the suggested ANOVA approaches: one-way ANOVA with Tukey’s HSD for strain comparisons (Figure 3), and two-way ANOVA (including the Genotype × Training interaction term) with Sidak's post-hoc test for learning assays (Figure 4). The updated statistical methods are now described in the Statistical Analysis section and figure legends.

The revised "Statistical Analysis" section is as follows (line 512-524):

"To ensure robust statistical analysis, each experimental group included at least 100 male flies (naïve, sexually experienced, or singly reared). Internal controls were incorporated into every experiment as recommended by Bretman et al. (2011) (Bretman et al., 2011). Normality of the mating duration data was confirmed using the Kolmogorov-Smirnov test (p>0.05). For group comparisons, two-sided Student’s t-tests were applied to calculate significance levels (****p<0.0001, ***p<0.001, **p<0.01, * p<0.05), Comparisons among three or more groups were performed using one-way ANOVA with Tukey’s HSD post-hoc tests. Two-factor experimental comparisons were performed using two-way ANOVA followed by Sidak’s post-hoc test. while estimation statistics (Claridge-Chang and Assam, 2016) were used to visualize effect sizes, mean differences, and precision, avoiding reliance solely on null hypothesis testing. All analyses, including data plotting, were performed using GraphPad Prism software."

(2) To decouple motor defects from courtship deficits in $w^{1118}$ and $y^1$ mutants, please use your tracking data to calculate and present a "General Locomotion" metric (e.g., average velocity or total distance traveled in the absence of a female).

We thank the reviewer for this suggestion. We have now added Figure S4 showing basal locomotor velocity of single-housed males across all four strains. As expected, w^1118 and y^1 mutants move more slowly than wild-type controls.

These data reveal that both motor and sensory factors contribute to the observed courtship phenotypes. Reduced basal locomotion likely limits the males' ability to approach and follow females. However, this generalized hypoactivity is compounded by strain-specific sensory deficits: w1118 males suffer visual impairment that compromises female detection (Krstic et al., 2013), while y1 males display altered cuticular hydrocarbons that disrupt pheromonal communication (Drapeau et al., 2006). These sensory defects impair courtship initiation and female recognition independent of locomotor capacity. Thus, the reduced CI, CL, and MD in these mutants reflect the combined effects of slower movement and courtship-specific sensory impairments, rather than motor defects alone.

We have revised the manuscript to incorporate the velocity data and clarify this interpretation (line 219-226):

"Notably, reduced basal locomotor activity in w1118 and y1 mutants has been well documented in previous studies, independent of courtship behavior (Drapeau et al., 2006; Krstic et al., 2013). Consistent with these reports, our tracking data show that single-housed w1118 and y1 males exhibit lower average velocity than Canton-S and Oregon-R controls (Fig. S4A). These general locomotor differences are insufficient to fully explain the observed courtship and mating timing phenotypes, indicating that additional courtship‑related processes contribute to the observed behavioral differences."

(3) Please expand the discussion or provide a small comparative dataset contrasting DrosoMating with established tools like JAABA. Explain the specific advantages of your pipeline (e.g., cost, simplicity, focus on CL/MD) to justify its adoption over these alternatives.

We sincerely appreciate the reviewer’s insightful comments on benchmarking DrosoMating against existing tools for automated courtship behavior analysis. We fully agree that direct comparisons are essential to clarify the unique advantages and ideal use cases of DrosoMating. We have extensively revised the manuscript by adding comparative experiments, a new figure (Figure 5) and comparative table (Table 1), and expanded descriptions, as detailed below.

(1) Direct comparison with conventional tracking tools

We systematically tested Ctrax and FlyTracker on the same low-quality, high-throughput mating videos used for DrosoMating. The corresponding results are presented in Figure 5.

Ctrax showed severe segmentation instability, including over-segmentation, under-segmentation, and fragmented trajectories, and failed to stably detect two flies.

FlyTracker was able to generate background models and showed partially improved segmentation after threshold adjustment, but frequently crashed during feature computation and could not complete the end-to-end analysis pipeline.

These results confirm that tracking-based tools are vulnerable to low contrast, chamber artifacts, and prolonged male–female overlap, whereas DrosoMating bypasses pose tracking and directly detects mating states.

The revised manuscript text is as follows (line 265-342):

" DrosoMating is more compatible with low-quality mating videos than conventional tracking-based pipelines

To evaluate whether conventional fly-tracking pipelines could be used as alternative tools for extracting mating-duration metrics, we tested Ctrax and FlyTracker on representative low-quality single-chamber mating videos recorded under our standard high-throughput conditions (Fig. 5). Ctrax is designed to estimate the position and orientation of multiple walking flies while maintaining individual identities over time (Branson et al., 2009) (https://ctrax.sourceforge.net/), whereas FlyTracker aims to track fly pose, including position, orientation, body size, wing and leg positions, and to generate trajectory- and feature-based outputs for downstream behavior analysis (Eyjolfsdottir et al., 2014). Because these programs were not readily compatible with our full high-throughput behavioral recording setup, we first cropped the original videos and tested single-chamber videos containing one male and one female fly.

Using Ctrax, we observed that target detection was highly variable even within single-chamber videos. In representative frames, Ctrax could occasionally identify the two flies correctly (Fig. 5A). However, imperfect segmentation was frequently observed. In some frames, parts of the fly body were detected as additional targets, resulting in over-segmentation and an apparent increase in the number of detected flies (Fig. 5B, C). Conversely, when the male and female were close to each other or physically overlapped, the two animals were sometimes detected as a single target (Fig. 5D). These examples indicate that Ctrax detection was sensitive to the low contrast and overlapping fly bodies present in our mating videos. We then adjusted the Ctrax detection threshold to improve segmentation quality. Although threshold optimization improved detection in some frames, abnormal detections remained evident across randomly sampled frames (Fig. 5E). For example, some frames still showed incorrect target numbers, and a severe segmentation failure was observed in the lower-right example, where the detected objects did not correspond to two clearly separable flies. Thus, even after parameter optimization, Ctrax did not consistently maintain the expected two-target detection state in single-chamber mating videos.

The instability of Ctrax detection was also reflected in the trajectory output. In the first 500 frames, the generated trajectories were fragmented into multiple colored track segments rather than two continuous trajectories corresponding to the male and female (Fig. 5F). In addition, some trajectories extended outside the chamber boundary, indicating tracking errors and identity instability. Consistently, frame-by-frame quantification of detected target number showed that the detected object count did not remain stable at the expected value of two flies per chamber (Fig. 5G). Because Ctrax failed to maintain stable two-fly detection and continuous trajectories under these video conditions, its output could not be reliably used for downstream extraction of copulation latency or mating duration.

We next tested FlyTracker on the same type of cropped single-chamber mating videos. FlyTracker was developed to track multiple flies by estimating body position, orientation, size, wing and leg positions, and by maintaining fly identities across video frames; it also outputs per-frame features such as velocity, facing angle, and wing-angle-related measurements for downstream behavioral analysis (Eyjolfsdottir et al., 2014). In our videos, FlyTracker was able to generate a background model, indicating that the program could recognize the overall imaging field and chamber background (Fig. 5H). However, during calibration and segmentation, the default threshold setting produced inconsistent detection results (Fig. 5I). Although some frames were segmented relatively well under the default threshold (Fig. 5J), these successful examples were not representative of the overall tracking process, and the program frequently terminated with runtime errors during tracking or downstream feature computation. To improve detection stability, we lowered the segmentation threshold. Under this adjusted setting, FlyTracker produced more complete fly masks in representative frames (Fig. 5K), and the diagnostic output showed that two flies were detected in many sampled frames (Fig. 5L). Nevertheless, detection remained unstable in some sampled frames, including frames in which zero or one fly was detected despite the expected two flies per chamber (Fig. 5L). The full pipeline ultimately failed during feature computation, producing a runtime error before complete tracking and feature outputs could be generated (Fig. 5M). Therefore, even after threshold adjustment, FlyTracker could not provide a stable end-to-end workflow for extracting mating-duration metrics from these low-quality mating videos.

This failure mode is relevant because FlyTracker depends on stable segmentation, identity maintenance, and per-frame feature extraction. In our assay videos, low contrast, chamber-edge artifacts, and prolonged male–female overlap during copulation interfered with these requirements. As a result, FlyTracker could occasionally identify the flies in individual frames, but it did not reliably complete the full analysis pipeline required for downstream behavioral quantification. This limitation is especially important for workflows such as JAABA, which use manually labeled examples to train behavior classifiers but still depend on upstream tracking-derived features. Thus, for our low-quality high-throughput mating recordings, FlyTracker-based analysis was substantially less practical than DrosoMating, which directly outputs mating-related timing metrics without requiring continuous high-fidelity two-fly pose tracking."

(2) Clarification of ideal use cases

We revised the Discussion to emphasize that DrosoMating is not intended to replace general-purpose pose or tracking tools (e.g., FlyTracker, Ctrax) that provide fine-grained behavioral data.

Instead, it offers a specialized, robust, and high-throughput workflow for scenarios requiring efficient extraction of mating timing metrics from low-quality videos, where conventional pipelines often fail.

These revisions greatly improve the clarity and positioning of DrosoMating. We thank the reviewer for this valuable suggestion.

The revised manuscript text is as follows (line 578-620):

“Comparative Evaluation with Existing Courtship Analysis Tools

A major advantage of DrosoMating is its compatibility with low-quality, high-throughput mating videos. Conventional tracking-based tools such as Ctrax and FlyTracker are powerful for trajectory- and pose-based behavioral analysis, but they generally require stable object segmentation, identity maintenance, and reliable feature extraction across frames. Ctrax was designed to estimate the positions and orientations of multiple walking flies while maintaining their identities, whereas FlyTracker aims to track detailed fly pose and generate per-frame behavioural features. Under our recording conditions, these requirements were difficult to satisfy because the videos were low contrast and the male and female frequently overlapped during copulation.

In our tests, both tools showed limited compatibility with these videos. Even after cropping to single-chamber videos and adjusting detection parameters, Ctrax produced unstable target numbers, fragmented trajectories, and tracking errors. FlyTracker could generate a background model and occasionally segment flies successfully, but detection remained unstable and the full pipeline failed during feature computation. These issues are particularly relevant for copulation latency and mating duration analysis: during copulation, the male and female remain physically coupled for a long period, which makes identity-based tracking difficult. For these timing metrics, it is more important to robustly detect the onset and offset of the mating state than to reconstruct detailed individual trajectories. DrosoMating was purpose-built to address these specific challenges. It operates reliably on lower-quality video streams, requires no complex pre-processing or manual ROI definition, and is optimized for high-throughput multi-chamber analysis. While it does not offer the same level of pose or kinematic detail as other tools, it provides a unique solution for laboratories seeking a simple, fast, and robust pipeline to quantify core reproductive timing metrics—copulation latency (CL), courtship index (CI), and mating duration (MD)—without the overhead of more complex systems (Table 1).

Although we directly tested only Ctrax and FlyTracker, this limitation may also affect workflows that depend on upstream tracking-derived features. For example, JAABA uses tracking-derived features to train behavior classifiers, and DANCE, a recent Drosophila aggression and courtship pipeline, uses JAABA-based classifiers and lists FlyTracker and JAABA as required software. Thus, our conclusion is not that these tools are generally unsuitable for Drosophila behavior analysis, but that DrosoMating provides a more practical workflow for low-quality, high-throughput videos focused specifically on mating timing.

While DrosoMating currently prioritizes mating timing metrics over discrete behavioral classification, its modular architecture provides a foundation for future integration with behavior classifiers such as JAABA. Laboratories requiring granular behavioral elements—such as wing extension or circling—would benefit from an extended pipeline that exports per-frame kinematic features for downstream classifier training. Validating this integration represents a promising future direction to broaden the tool’s utility beyond core reproductive timing assays.”

We have added this clarification to the Table 1 legend to avoid overgeneralizing the limitations of Ctrax and FlyTracker (line 786-790):

"DrosoMating was designed to extract mating-related timing metrics from high-throughput videos without requiring continuous two-fly identity tracking. Ctrax and FlyTracker were tested on cropped single-chamber videos from the same recording setup. Their limitations described here refer specifically to these low-quality mating videos and should not be interpreted as general limitations of the tools."

(4) Complement the aggregate bar plots with behavioral ethograms or raster plots for representative individual flies. Color-code these plots for specific states (resting, following, courting, copulating) to visually demonstrate the system's temporal resolution and detection accuracy.

We thank the reviewer for this valuable suggestion. To address this comment, we have added a new supplementary figure (Figure S3) that presents behavioral ethograms for individual flies, directly complementing the aggregate bar plots in the main text. The figure displays the full temporal progression of courtship and copulation behaviors for all wells that exhibited successful mating. As requested, behaviors are color-coded (orange: courting, red: copulating) to clearly delineate different states. These ethograms visually demonstrate the system’s ability to resolve behavioral transitions with high temporal precision, confirming the accuracy of our automated detection of courtship initiation, duration, and copulation events. This addition provides critical individual-level validation that supports the aggregate statistical results presented in the main text.

(5) Standardize the use of estimation statistics (DBMs). If used in Figure 3, they should also be applied to Figure 4, with appropriate statistical comparisons between groups.

We thank the reviewer for this suggestion. We have standardized the statistical reporting format across all figure legends, with each legend explicitly stating the test used, the post hoc method (where applicable), and significance thresholds.

Our approach is as follows:

Fig. 2 and Fig.3D-I (two-group comparison, manual vs. automated scoring): DBM + Student's t-test.

Fig. 3A-C and Fig. S1B-C, E-F (multi-group comparison, 4 strains or 2 rearing conditions): One-way ANOVA + Tukey's HSD.

Fig. 4B-D and F (two-factor design, genotype × training): Two-way ANOVA + Sidak's post hoc.

We have also added a sentence to the Methods clarifying that estimation statistics (DBM) are used for single two-group contrasts, while ANOVA-based approaches are used for multi-group or multi-factor designs. The statistical methods are now consistently documented in both the Methods section and the corresponding figure legends. We hope this clarification addresses the reviewer's concern.

The revised "Statistical Analysis" section is as follows (line 512-524):

"To ensure robust statistical analysis, each experimental group included at least 100 male flies (naïve, sexually experienced, or singly reared). Internal controls were incorporated into every experiment as recommended by Bretman et al. (2011) (Bretman et al., 2011). Normality of the mating duration data was confirmed using the Kolmogorov-Smirnov test (p>0.05). For group comparisons, two-sided Student’s t-tests were applied to calculate significance levels (****p<0.0001, ***p<0.001, **p<0.01, * p<0.05), Comparisons among three or more groups were performed using one-way ANOVA with Tukey’s HSD post-hoc tests. Two-factor experimental comparisons were performed using two-way ANOVA followed by Sidak’s post-hoc test. while estimation statistics (Claridge-Chang and Assam, 2016) were used to visualize effect sizes, mean differences, and precision, avoiding reliance solely on null hypothesis testing. All analyses, including data plotting, were performed using GraphPad Prism software."

(6) Fix inconsistent labeling (e.g., "MB-247" vs. "MB247") and redundant axis labels.

Thank you for pointing this out. We have now standardized all labels to "MB247-GAL4" throughout the text and figures.

We have also removed redundant axis labels from multi-panel figures. And redundant DBM and metric labels have been streamlined in Figures 2 and 4. All statistical tests and metric definitions are now fully described in the corresponding figure legends rather than being repeated on each sub-panel.

(7) Significantly increase the size of data panels to make individual data points and error bars legible.

Thank you for this suggestion. We have standardized the figure formatting and simplified the panels by removing redundant axis labels and consolidating descriptive details into the figure legends. We believe the current panel sizes, combined with these clarifications, provide sufficient legibility for both data points and error bars in the final high-resolution PDF.

Minor Corrections:

(1) Abstract: Rephrase "lack of certain timing-related behavioral repertoires" (Line 108) to "lack of precise quantification for temporal parameters of post-copulatory behavior."

Thank you for the suggested rephrasing. We have updated the sentence accordingly.

(2) Define the physical parameter "s" (e.g., is it a pixel threshold?) to ensure reproducibility.

Thank you for this important suggestion. We have now explicitly defined the parameter s in the Methods section. Briefly, s is the grayscale intensity threshold (0–255, 8-bit) used for binary segmentation of flies from the background. It is automatically calculated as the maximum grayscale value of three user-selected reference flies plus an offset of 28, and can be manually adjusted to accommodate varying illumination conditions.

The revised manuscript text is as follows (line 499-503):

“Note on the s value: The parameter s represents the grayscale intensity threshold (range: 0–255 for 8-bit images) used for binary segmentation of flies from the background. It is automatically calculated as the maximum grayscale value at the three reference fly positions plus an offset of 28, and can be manually adjusted to accommodate varying illumination conditions.”

(3) Figure 1 Legend: Clarify the "clockwise selection" of pillars and their relation to well numbering.

Thank you for this suggestion. We have revised the Figure 1 legend to clarify that:

(1) The four pillars are selected in clockwise order starting from the top-left corner to define the chamber corners for perspective transformation (homography), which corrects for camera angle and standardizes the field of view.

(2) Well numbering is independent of pillar selection order. After automated perspective correction, wells are numbered sequentially in a left-to-right, top-to-bottom order (1–36) based on the standardized chamber layout.

The updated Figure 1 legend now reads:

"Columns (pillars) should be selected in clockwise order starting from the top-left corner to define the chamber boundaries for perspective transformation (Fig. 1A, lower). Well numbering (1–36) follows a left-to-right, top-to-bottom sequence after perspective correction and is independent of pillar selection order."

(4) Add the citation for Eastwood and Burnet (1977) regarding Courtship Latency.

Thank you for pointing this out. We have added the citation Eastwood and Burnet (1977) to the Introduction where Courtship Latency is first defined

(5) Select one term ("Mating Duration" or "Copulation Duration") and use it consistently throughout the text and figures.

Thank you for this suggestion. We have now standardized the terminology throughout the manuscript and figures. "Mating Duration" (MD) is used consistently in all instances where "Copulation Duration" previously appeared. The abbreviation MD has been retained for consistency with existing figure labels

References

Branson, K., Robie, A.A., Bender, J., Perona, P., & Dickinson, M.H. (2009). High-throughput ethomics in large groups of Drosophila. Nature Methods, 6(6), 451–458.

Claridge-Chang, A., & Assam, P.N. (2016). Estimation statistics should replace significance testing. Nature Methods, 13(2), 108–109.

Drapeau, M.D., Cyran, S.A., Viering, M.M., Geyer, P.K., & Long, A.D. (2006). A cis-regulatory sequence within the yellow locus of Drosophila melanogaster required for normal male mating success. Genetics, 172(2), 1009–1030.

Eastwood, L., & Burnet, B. (1977). Courtship latency in male Drosophila melanogaster. Behavior Genetics, 7(3), 359–372.

Eyjolfsdottir, E., Branson, S., Burgos-Artizzu, X.P., Hoopfer, E.D., Schor, J., Anderson, D.J., & Perona, P. (2014). Detecting social actions of fruit flies. Lecture Notes in Computer Science, 8692, 772–787.

Gil-Martí, B., Barredo, C.G., Pina-Flores, S., Poza-Rodriguez, A., Treves, G., Rodriguez-Navas, C., Camacho, L., Pérez-Serna, A., Jimenez, I., Brazales, L., Fernandez, J., & Martin, F.A. (2023). A simplified courtship conditioning protocol to test learning and memory in Drosophila. STAR Protocols, 4(1), 101572.

Greenspan, R.J., & Ferveur, J.F. (2000). Courtship in Drosophila. Annual Review of Genetics, 34, 205–232.

Hall, J.C. (1994). The mating of a fly. Science, 264(5163), 1702–1714.

Kabra, M., Robie, A.A., Rivera-Alba, M., Branson, S., & Branson, K. (2013). JAABA: interactive machine learning for automatic annotation of animal behavior. Nature Methods, 10(1), 64–67.

Kitamoto, T. (2001). Conditional modification of behavior in Drosophila by targeted expression of a temperature-sensitive shibire allele in defined neurons. Journal of Neurobiology, 47(2), 81–92.

Krstic, D., Boll, W., & Noll, M. (2013). Influence of the White locus on the courtship behavior of Drosophila males. PLoS ONE, 8(9), e77904.

Levin, L.R., Han, P.L., Hwang, P.M., Feinstein, P.G., Davis, R.L., & Reed, R.R. (1992). The Drosophila learning and memory gene rutabaga encodes a Ca2+/calmodulin-responsive adenylyl cyclase. Cell, 68(3), 479–489.

Pavlou, H.J., & Goodwin, S.F. (2013). Courtship behavior in Drosophila melanogaster: towards a ‘courtship connectome’. Current Opinion in Neurobiology, 23(1), 76–83.

Yapici, N., Kim, Y.J., Ribeiro, C., & Dickson, B.J. (2008). A receptor that mediates the post-mating switch in Drosophila reproductive behaviour. Nature, 451(7176), 33–37.

Version 2

Abstract

Male Drosophila courtship behavior is a key model for studying temporal decision-making. While courtship index (CI) is widely used to quantify mating activity, other timing metrics like courtship latency, copulation latency (CL), and mating duration (MD) remain understudied. Traditional methods for quantifying these behaviors are often labor-intensive and prone to human error. In this study, we present a protocol combining a modular chamber system and automated software (DrosoMating) to quantify 6 key timing metrics during male courtship. Our image-based video analysis enables precise identification of courtship and copulation events, as well as quantification of their timing and duration under controlled conditions. Validation shows <0.05% error rates and 98–99% agreement with manual scoring for CL, CI, and MD. The protocol supports genetic and neural circuit manipulations, detecting subtle genetic, social, and environmental behavioral variations. By minimizing manual effort and standardizing data collection, this approach facilitates scalable, reproducible studies on adaptive trade-offs, learning, and neural mechanisms in mating behavior. This method streamlines timing analysis in male courtship, offering reproducible metrics for behavioral genetics.

Introduction

Time perception is critical for survival, enabling organisms to estimate intervals between events and execute adaptive behaviors. Male Drosophila courtship behavior, a genetically well-characterized model (Singh and Singh, 2016), provides a robust framework for studying temporal control in action sequences. A widely used metric to quantify mating activity is the ‘Courtship Index (CI)’, defined as the percentage of time a male engages in courtship behaviors (e.g., following, wing vibration, abdominal curling) toward a female during a standardized observation window. CI reflects both the intensity and persistence of courtship, serving as a sensitive readout for dissecting genetic, environmental, or experimental perturbations on mating behavior, as demonstrated in studies investigating genetic mutations, environmental perturbations, and neuronal manipulations (Benton, 2015; Greenspan and Ferveur, 2000a; Kim, 2009; Pavlou and Goodwin, 2013; Yamamoto and Koganezawa, 2013).

Despite extensive research on courtship displays (Chen et al., 2024a), other critical metrics remain poorly characterized, especially as their representative characteristics as timing behaviors. ‘Courtship latency’ (time until a male initiates courtship after detecting a female) may reflect motivational state or sensory processing deficits but is rarely quantified in terms of genetics (Eastwood and Burnet, 1977a). ‘Copulation Latency (CL)’ (time from courtship initiation to successful mating) and ‘Mating Duration (MD)’ (length of mating itself) are similarly underexplored, though both likely impact reproductive success. Prolonged MD enhances paternity assurance by maximizing sperm transfer and suppressing female remating, but it simultaneously increases predation risks and energy expenditure (Bretman et al., 2009). Conversely, reduced MD may decrease a male’s paternity share due to incomplete sperm transfer or female resistance, yet it reduces exposure to ecological threats (e.g., predators, rivals) and conserves metabolic resources (Lee et al., 2023a). These traits may be influenced by cryptic factors such as male age, rival presence, or environmental stressors (e.g., temperature), yet their genetic and neural bases are unclear. While studies focus on stereotyped courtship actions, these “temporal metrics” offer insights into decision-making, persistence, and adaptive trade-offs—areas ripe for investigation (Huang et al., 2024; Kim et al., 2012a, 2013a, 2016, 2024; Lee et al., 2022; D. Sun et al., 2024; Y. Sun et al., 2024; Wong et al., 2019; T. Zhang et al., 2024b, 2024a; X. Zhang et al., 2024).

Numerous software tools have been developed to measure male mating behavior in Drosophila, yet many laboratories still rely on manual methods. Traditional approaches, such as the “courtship wheel”, have been widely used to assess male courtship activity (Hall, 1994; Neckameyer and Bhatt, 2016; Vosshall, 2007). However, quantifying the full repertoire of mating behavior metrics has historically been labor-intensive (Ejima and Griffith, 2007; Sokolowski, 2001). Recent advances in automation have addressed these challenges. For instance, Dankert et al. (2009) introduced a machine vision system to quantify male-female interactions, though its application has been more focused on studying male-male aggression (Dankert et al., 2009a). With the rapid development of machine learning and image processing, tools such as ‘Ctrax’ (Branson et al., 2009), ‘idTracker’ (Pérez-Escudero et al., 2014), Cadabra (Stern et al., 2015), ‘FlyTracker’ and MATLAB combined with ‘JAABA’ (Barwell et al., 2021; Kabra et al., 2013) have enabled precise quantification of male-female mating behavior. Recently, Chen et al. (2024) introduced an image-processing-based system that addresses this gap by automatically recognizing multiple courtship elements with high accuracy (>97%) and robust identity tracking even during complete fly overlap(Chen et al., 2024b). Despite these advancements, recent efforts to simplify courtship conditioning protocols (Gil-Martí et al., 2023a) still lack precise quantification for temporal parameters of post-copulatory behavior. Consequently, there is a need for more streamlined protocols and software to accurately measure the timing aspects of male mating behavior in Drosophila.

Beyond Drosophila-specific tools, general behavioral video analysis paradigms have been developed to track the movement and behavior of various organisms, including fruit flies. These methods often focus on tracking the position of single or multiple flies (Iyengar et al., 2012; Qu et al., 2022; Simon et al., 2011) or using artificial markers (Gal et al., 2020). Additionally, programs like ‘ToxTrac’ (Rodriguez et al., 2017), ‘UMATracker’ (Yamanaka and Takeuchi, 2018), ‘FIMTrack’ (Risse et al., 2017, 2014), and ‘ilastik’ (Berg et al., 2019) have been employed to analyze body posture, behavior, location, and movement trajectories in diverse species, including flies, bees, mice, and zebrafish. However, these tools often fall short in capturing the nuanced courtship behaviors of Drosophila, as motion data alone may not sufficiently represent complex interactions.

To address these limitations, we have developed a specialized software suite called ‘DrosoMating’ for analyzing and quantifying timing behaviors within male Drosophila courtship. By establishing 6 key indicators related to male courtship, we provide a robust framework for quantifying male mating behavior metrics under controlled conditions. This approach significantly reduces the time and errors associated with manual identification. Through iterative optimization, our system achieves an error rate of less than 0.05% compared to manual scoring. We believe this toolset offers a novel and efficient method for studying timing in male courtship behavior, providing valuable data and theoretical insights while streamlining the analysis process.

Results

Video-Processing Solutions for Precise Drosophila Identification in Multi-Chamber Setups

Accurate identification of Drosophila individuals is essential for behavioral quantification given their small size. DrosoMating addresses this through three integrated video-processing features that enhance recognition fidelity in multi-chamber environments.

First, an effective recognition region function isolates individual wells as discrete analysis zones (Fig. 1A bottom), eliminating interference from adjacent chambers and background artifacts. Second, an image learning function adapts to phenotypic variability across strains. Operators designate three sample flies (Fig. 1B), enabling the algorithm to internalize strain-specific morphological signatures for improved detection. Third, a manual threshold adjustment function permits post-recognition optimization. Users visually validate automated identifications and dynamically refine sensitivity parameters (Fig. 1C), correcting false positives/negatives. Collectively, these features ensure robust tracking accuracy across diverse experimental setups and genetic backgrounds.

Usage of DrosoMating and CAD Parameters of the Chamber

(A) Typing input speed and input board numbers before selecting analyzing area (upper). Schematic representation of the region selection process (lower). Columns should be selected in a clockwise direction starting from the top left corner to define the chamber boundaries for perspective transformation (Fig. 1A, lower). Well numbering (1–36) follows a left-to-right, top-to-bottom sequence after perspective correction and is independent of pillar selection order. (B) Selection of reference flies for tracking accuracy. Any three reference flies were chosen for the analysis. (C) Use the X-axis, Y-axis, and threshold regulator to adjust the selection area to fit each well. Click “Enter” to run the program. Output file will be in the video folder: [Video Name]_output_[board numbers].csv (D) CAD (computer-aided design) exploded 3D view of the chamber instruction. (E) CAD parametric data of the chamber (dimensions in mm). (F) 3D diagram of the completed chamber

Development and Validation of DrosoMating for Automated Drosophila Mating Behavior Analysis

To address the labor-intensive nature, high error rates, and limitations in quantifying complex behavioral metrics during manual observation of Drosophila mating, we developed the software DrosoMating. This tool automates analysis of mating videos recorded in a customized chamber. DrosoMating processes videos to generate a CSV file containing 6 key behavioral metrics: courtship duration, MD, CI, courtship start time, mating start time, and mating end time. While these metrics are individually straightforward, they can be combined to quantify diverse aspects of mating behavior metrics.

DrosoMating enables quantification of specific mating behaviors based on research needs. We focused on three core indices:

Courtship Index (CI): Indicates male courtship investment per unit time. Higher CI values (reflecting longer durations of behaviors like wing vibration or female chasing) suggest stronger male courtship motivation or higher female attractiveness (Greenspan and Ferveur, 2000b).

Copulation Latency (CL): Measures the time from courtship initiation to copulation start. Shorter CL indicates efficient male courtship or high female receptivity. Longer CL may reflect male courtship deficits (Eastwood and Burnet, 1977a), intense female rejection, or environmental stressors (Greenspan and Ferveur, 2000b; Singh and Singh, 2014).

Mating Duration (MD): Represents the copulation period length. Longer MD may enhance sperm competition but increases predation risk and energy costs, illustrating an evolutionary trade-off between reproduction and survival (Greenspan and Ferveur, 2000b).

Together, CI, CL, and MD provide a robust framework for characterizing Drosophila behavioral states during mating. Accurate quantification of these metrics is essential.

Because overlapping behaviors occur frequently during courtship and mating, DrosoMating handles short and prolonged overlaps differently. When flies overlap for ≤15 frames, centroid velocity continuity is used to assign identity; if overlap >15 frames, the region is flagged ‘occluded’ and excluded from CI calculation. Occluded frames are excluded from CI calculation but retained for mating-state detection via merged-contour analysis, ensuring continuous MD measurement through copulation.

We validated DrosoMating by comparing its output for CI, CL, and MD against values from repeated, corrected manual measurements. Differences between software-generated data and manual data were consistently within 10 seconds for all three metrics, with no statistically significant differences (Fig. 2A-C). Manual quantification of these metrics requires several hours of effort and is prone to substantial human error. In contrast, DrosoMating automates the analysis. After a few minutes of initial setup and configuration, the software generates precise behavioral metrics from mating videos (Fig. S3).

Comparison of the Accuracy of DrosoMating with Manual Scoring

(A) Copulation latency (CL) of Canton-S males was analyzed using DrosoMating and compared with manual recordings. For this figure, DBMs represent the ‘difference between means’ for the evaluation of estimation statistics. Asterisks represent significant differences, as revealed by the Student’s t test (* p<0.05, ** p<0.01, *** p<0.001). Differences between methods were assessed using two-sided Student’s t-test (ns, not significant). See the Background for a detailed description of the CL assays used in this study. (B) Courtship index (CI) of Canton-S was analyzed by DrosoMating and manual recording. CI was measured by [courtship time/ (mating time-courtship time) *100%]. For detailed methods, see the BACKGROUND for a detailed description of the CI assays used in this study. (C) Mating Duration (MD) of Canton-S was analyzed by DrosoMating and manual recording. See the BACKGROUND for a detailed description of the MD assays used in this study.

Validation of DrosoMating Across Diverse Drosophila Strains and Behavioral Paradigms

DrosoMating was adapted to quantify mating behavior metrics across diverse Drosophila strains and experimental conditions. To evaluate its universality and accuracy, mating interactions were analyzed between virgin females and males of four strains: Canton-S (wild-type), Oregon-R (wild-type), w1118 (white eye mutant), and y1 (yellow body color mutant). These lines have been extensively employed as standard inbred strains in research contexts for a prolonged period. Wild-type strains (Canton-S and Oregon-R) exhibited significantly higher CI, CL, and MD compared to w1118 and y1 mutants (Fig. 3A–C). This quantitative assessment revealed markedly reduced activity in mutant males across all three behavioral metrics, consistent with known phenotypic limitations: w1118 males displayed restricted courtship due to visual impairment (Krstic et al., 2013), while y1 males demonstrated suboptimal mating performance, aligning with prior reports of courtship deficits (Drapeau et al., 2006).

Behavioral assays of Drosophila males under different rearing conditions.

(A) CL of Canton-S, w1118, y1, and Oregon-R males. (B) CI of Canton-S, w1118, y1, and Oregon-R males. (C) MD of Canton-S, w1118, y1, and Oregon-R males. For detailed methods, see the background for a detailed description of the mating assays used in this study. Violin plots show the distribution of copulation latency (CL, in minutes) for Canton-S, w¹¹¹8, y¹, and Oregon-R males. Sample sizes are indicated below each group. Statistical significance was assessed using one-way ANOVA followed by Tukey’s post hoc test for multiple comparisons for Fig. 3A-C. ****p < 0.0001; ns, not significant. (D) CL of group-housed and single-reared Canton-S males. For Fig. 3D-I, DBMs represent the ‘difference between means’ for the evaluation of estimation statistics. Asterisks represent significant differences, as revealed by the Student’s t test (* p<0.05, ** p<0.01, *** p<0.001). Differences between methods were assessed using two-sided Student’s t-test (ns, not significant). (E) CI of group-housed and single-reared Canton-S males. (F) LMD assays of Canton-S males. In the MD assays, white data points denote males that were group-reared (or sexually naïve), whereas blue data points signify males that were singly reared. The dot plots represent the MD of each male fly. The mean value and standard error are labeled within the dot plot. (G) CL of naive and sexually experienced Canton-S males. (H) CI of naive and sexually experienced Canton-S males. (I) SMD assays of Canton-S males. White data points represent sexually-naïve males and pink data points represent sexually-experienced ones.

Notably, reduced basal locomotor activity in w¹¹¹8 and y¹ mutants has been well documented in previous studies, independent of courtship behavior (Drapeau et al., 2006; Krstic et al., 2013). Consistent with these reports, our tracking data show that single-housed w1118 and y1 males exhibit lower average velocity than Canton-S and Oregon-R controls (Fig. S4A). These general locomotor differences are insufficient to fully explain the observed courtship and mating timing phenotypes, indicating that additional courtship_related processes contribute to the observed behavioral differences.

The software further validated established behavioral paradigms—Longer Mating Duration (LMD) (Kim et al., 2013a, 2012a) and Shorter Mating Duration (SMD) (Lee et al., 2023a)—using Canton-S males subjected to distinct pre-experimental conditioning. DrosoMating accurately detected expected differences in CL, CI, and MD between LMD (group vs. single housing) and SMD (sexually naive vs. sexually experienced) conditions (Fig. 3D–I), confirming its precision in resolving complex, paradigm-specific mating dynamics.

Quantification of Learning-Memory Dynamics and Neural Circuit Integration Using DrosoMating

DrosoMating facilitates the investigation of learning and memory consolidation in Drosophila through quantifiable behavioral metrics. The Learning Index (LI) was calculated as LI = (CI_naive − CI_trained) / CI_naive × 100%. (Fig. 4A), This metric provides a standardized measure of experience-dependent behavioral modulation. This transformation of abstract behavior into discrete parameters enables rigorous analysis of neural and behavioral plasticity.

behavioral assays of learning in Drosophila which combined with temperature-dependent temporal activation/inhibition methods.

(A) Learning Index (LI) calculation and experimental design for temperature-dependent activation/inhibition of neural circuits in adult Drosophila. This figure illustrates the experimental paradigm for calculating the LI using CI measurements from sham and trained group. The protocol involves rearing flies at 20°C for four days followed by a one-day exposure to 29°C to activate/inhibit temperature-dependent, genetically-encoded neuronal modulators (for example, shits as inhibitor or TrpA1 as activator). (B) CI of training and sham training MB247-GAL4/UAS-GFP, MB247-GAL4/UAS-TrpA1 and MB247-GAL4/UAS-shits flies. Sample sizes (n) are indicated below each group. Statistical significance was assessed using two-way ANOVA for Fig. 4 B,C,D and F, followed by Sidak’s post hoc test for pairwise comparisons between sham and trained groups within each genotype. ****p < 0.0001; ns, not significant. (C) CI of different training days (D1: 1 day, D2: 2 days, D3: 3 days) and sham trained Canton-S flies. (D) CI of different training days (D1: 1day, D2: 2 days, D3: 3 days) and sham training orbΔ/orbΔQ flies. (E) LI of Canton-S and orbΔ/orbΔQ males. For fig. 4E and G, DBMs represent the ‘difference between means’ for the evaluation of estimation statistics. Asterisks represent significant differences, as revealed by the Student’s t test (* p<0.05, ** p<0.01, *** p<0.001). Differences between methods were assessed using two-sided Student’s t-test (ns, not significant). (F) CI of training and sham training rut2080; UAS-rut, rut2080; GAL4OK107, rut2080; GAL4OK107/UAS-rut flies. (G) LI of rut2080; UAS-rut, rut2080; GAL4OK107, rut2080; GAL4OK107/UAS-rut flies.

The software integrates seamlessly with neural manipulation techniques. For acute temporal control, thermogenetic tools (UAS-shits (Kitamoto, 2001), UAS-TrpA1 (Kang et al., 2012)) permit precise activation or silencing of circuits during courtship (Fig. 4B and S1A). Targeted perturbation of mushroom body Kenyon cells, critical for memory processing, revealed immediate behavioral consequences (Fig. S1B–C), elucidating their role in behavioral adaptation.

Established courtship conditioning paradigms (Gil-Martí et al., 2023a; Griffith and Ejima, 2009a; Kamyshev et al., 1999; Reza et al., 2013; Wolf et al., 1998) are robustly replicated, validating DrosoMating’s fidelity in experience-driven behavior. (Fig. 4C). Mutant analyses revealed orb2 (Fig. 4D–E) and rut (Fig. 4F-G) dependence of experience-induced CI reduction, consistent with known memory pathways.

Chronic neural interventions are enabled via tub-GAL80ts-mediated adult-specific silencing or knockdown (Fig. S1D–F). This approach reveals enduring impacts of sustained circuit modulation on courtship plasticity and memory consolidation. Collectively, DrosoMating bridges behavioral quantification and mechanistic inquiry, advancing exploration of learning-memory interplay through precise, scalable experimental frameworks.

DrosoMating is more compatible with low-quality mating videos than conventional tracking-based pipelines

To evaluate whether conventional fly-tracking pipelines could be used as alternative tools for extracting mating-duration metrics, we tested Ctrax and FlyTracker on representative low-quality single-chamber mating videos recorded under our standard high-throughput conditions (Fig. 5). Ctrax is designed to estimate the position and orientation of multiple walking flies while maintaining individual identities over time (Branson et al., 2009) (https://ctrax.sourceforge.net/), whereas FlyTracker aims to track fly pose, including position, orientation, body size, wing and leg positions, and to generate trajectory- and feature-based outputs for downstream behavior analysis (Eyjolfsdottir et al., 2014). Because these programs were not readily compatible with our full high-throughput behavioral recording setup, we first cropped the original videos and tested single-chamber videos containing one male and one female fly.

Limitations of conventional tracking-based tools for low-quality Drosophila mating videos.

(A-D) Representative Ctrax detection frames from single-chamber mating videos, showing examples with different detected target numbers. (E) Additional randomly sampled Ctrax detection frames during tracking. (F) Representative Ctrax trajectory plot generated from the tracked video. (G) Frame-by-frame plot of the number of targets detected by Ctrax. (H) FlyTracker background model, background variance, and sample image used for detection. (I-K) Representative FlyTracker segmentation outputs under different threshold settings. (I) Segmentation output using the default threshold setting. (J) Example frame showing relatively successful fly segmentation under the default threshold setting. (K) Representative segmentation output after lowering the threshold to reduce tracking/feature-computation errors. (L) FlyTracker command-window output showing the number of flies detected in sampled frames. (M) Example FlyTracker error message during feature computation.

Using Ctrax, we observed that target detection was highly variable even within single-chamber videos. In representative frames, Ctrax could occasionally identify the two flies correctly (Fig. 5A). However, imperfect segmentation was frequently observed. In some frames, parts of the fly body were detected as additional targets, resulting in over-segmentation and an apparent increase in the number of detected flies (Fig. 5B,C). Conversely, when the male and female were close to each other or physically overlapped, the two animals were sometimes detected as a single target (Fig. 5D). These examples indicate that Ctrax detection was sensitive to the low contrast and overlapping fly bodies present in our mating videos. We then adjusted the Ctrax detection threshold to improve segmentation quality. Although threshold optimization improved detection in some frames, abnormal detections remained evident across randomly sampled frames (Fig. 5E). For example, some frames still showed incorrect target numbers, and a severe segmentation failure was observed in the lower-right example, where the detected objects did not correspond to two clearly separable flies. Thus, even after parameter optimization, Ctrax did not consistently maintain the expected two-target detection state in single-chamber mating videos.

The instability of Ctrax detection was also reflected in the trajectory output. In the first 500 frames, the generated trajectories were fragmented into multiple colored track segments rather than two continuous trajectories corresponding to the male and female (Fig. 5F). In addition, some trajectories extended outside the chamber boundary, indicating tracking errors and identity instability. Consistently, frame-by-frame quantification of detected target number showed that the detected object count did not remain stable at the expected value of two flies per chamber (Fig. 5G). Because Ctrax failed to maintain stable two-fly detection and continuous trajectories under these video conditions, its output could not be reliably used for downstream extraction of copulation latency or mating duration.

We next tested FlyTracker on the same type of cropped single-chamber mating videos. FlyTracker was developed to track multiple flies by estimating body position, orientation, size, wing and leg positions, and by maintaining fly identities across video frames; it also outputs per-frame features such as velocity, facing angle, and wing-angle-related measurements for downstream behavioral analysis(Eyjolfsdottir et al., 2014). In our videos, FlyTracker was able to generate a background model, indicating that the program could recognize the overall imaging field and chamber background (Fig. 5H). However, during calibration and segmentation, the default threshold setting produced inconsistent detection results (Fig. 5I). Although some frames were segmented relatively well under the default threshold (Fig. 5J), these successful examples were not representative of the overall tracking process, and the program frequently terminated with runtime errors during tracking or downstream feature computation. To improve detection stability, we lowered the segmentation threshold. Under this adjusted setting, FlyTracker produced more complete fly masks in representative frames (Fig. 5K), and the diagnostic output showed that two flies were detected in many sampled frames (Fig. 5L). Nevertheless, detection remained unstable in some sampled frames, including frames in which zero or one fly was detected despite the expected two flies per chamber (Fig. 5L). The full pipeline ultimately failed during feature computation, producing a runtime error before complete tracking and feature outputs could be generated (Fig. 5M). Therefore, even after threshold adjustment, FlyTracker could not provide a stable end-to-end workflow for extracting mating-duration metrics from these low-quality mating videos.

This failure mode is relevant because FlyTracker depends on stable segmentation, identity maintenance, and per-frame feature extraction. In our assay videos, low contrast, chamber-edge artifacts, and prolonged male–female overlap during copulation interfered with these requirements. As a result, FlyTracker could occasionally identify the flies in individual frames, but it did not reliably complete the full analysis pipeline required for downstream behavioral quantification. This limitation is especially important for workflows such as JAABA, which use manually labeled examples to train behavior classifiers but still depend on upstream tracking-derived features. Thus, for our low-quality high-throughput mating recordings, FlyTracker-based analysis was substantially less practical than DrosoMating, which directly outputs mating-related timing metrics without requiring continuous high-fidelity two-fly pose tracking.

Methods and Materials

Fly Stocks and Husbandry

Drosophila stocks were maintained on standard cornmeal-agar food at 25°C and 60-70% relative humidity, with a 12-hour light/dark cycle. When suppression of temperature-sensitive transgene activity was required (e.g., for UAS-shits or UAS-TrpA1 experiments), flies were reared at 19-20°C throughout development. Because developmental duration at 19-20°C is approximately twice that at 25°C, experimental scheduling was adjusted to accommodate the delayed eclosion. Transgene activation was achieved by transferring flies to 29°C (shits) or 28°C (TrpA1) for 30 mins prior to experiments. To eliminate confounding variables in behavioral assays, control and experimental groups were always raised under identical environmental conditions, including temperature, humidity, and light cycle.

Environmental Control System

The controlled-environment chamber is a dedicated fly room equipped to maintain standardized experimental conditions. It features a precision temperature control unit, which is a programmable apparatus for maintaining constant temperature with an accuracy of ±0.5°C. The chamber also includes automated humidity regulation through a humidifier or climate control system to stabilize relative humidity, typically within the range of 60–70%. Additionally, a programmable LED lighting system with full-spectrum arrays and an integrated timer is used to simulate circadian cycles, allowing for adjustable day/night duration and light intensity.

Mating Behavior Measurement System Setup

Behavioral observation arena: The behavioral observation arena was constructed from transparent acrylic plates. Each chamber measured approximately 1 × 1 × 0.2 cm unless otherwise specified. (Fig. 1D-E).

Sex-separation films: Non-permeable acrylic partitions (0.5 mm thickness) to isolate male and female cohorts prior to trials. Customized apparatus components can be fabricated by precision cutting of OHP (overhead projector) film sheets.

Uniform illumination system: Programmable LED panel array (full-spectrum, 6500K color temperature) positioned to optimize contrast for automated tracking software.

High-resolution imaging: OPPO A72 5G main camera (FHD/1080p resolution, 30 fps) mounted on an adjustable stabilizer for consistent video capture.

Mating Behavior Measurement Preparations

This protocol outlines the collection of male and virgin female Drosophila melanogaster and details the standardized mating behavior assay. To ensure experimental reproducibility, each cohort (genotype or condition) must include ≥100 individuals per sex, with final mating counts reaching at least 20 successful pairs (ideally 70-80) within a 2-hour observation window. The protocol spans developmental stages from egg to adult emergence, with adults requiring 5 days post-eclosion to attain sexual maturity. Flies are reared under controlled environmental conditions (25°C, 60-70% relative humidity, 12-hour light/dark cycle) to minimize physiological variability. To avoid confounding effects of anesthesia on mating behavior, CO2 exposure was strictly minimized during fly handling and sorting; brief, low-intensity exposure was reserved for initial virgin collection and before starting behavioral assays.

Preparation of experimental male flies (15–20 days): Raise flies of the desired genotype (wild-type, mutants, or crosses such as GAL4/UAS combinations) in bottles at the specified temperature (standard: 25°C; alternative: 19°C or 29°C as required) under a 12:12 h light:dark cycle. Maintain 2–3 bottles per genotype, each containing 25 females and 15 males, to yield 25–30 flies per day. Transfer flies to fresh food every 2–3 days to prevent overcrowding.

Post-eclosion handling: Collect newly eclosed males (0–8 h post-eclosion) under brief CO2 anesthesia. Isolate individual males in vials and age them for 5 days at 25°C (or 9 days at 18°C) to ensure complete brain maturation. Clear existing adults from culture vials daily using CO2 to ensure synchronized collection of newly eclosed males. Group males into cohorts of 40 and rear in vials for 5–9 days to optimize sexual maturity while preserving mating efficiency.

Large-scale virgin female collection for high-throughput genetic screening: The Df(1)Exel6234/Yhs-hid (Bloomington Stock Center #7708 with #8846) or the SPRattP/Yhs-hid strain (Bloomington Stock Center #84576 with #8846) were used in this study. Both strains carry a null mutation in the sex-peptide receptor (SPR) gene (Yapici et al., 2008), which disrupts post-mating female refractoriness, thereby enhancing receptivity to remating and heat-inducible hid expression eliminating male progeny and balancer heterozygotes. Maintain cultures below 20°C prior to egg collection to prevent premature hid-induced lethality in Yhs-hid containing stocks. Higher temperatures (>20°C) may activate leaky hs-hid expression, reducing male lethality efficiency during subsequent heat shock.

Place adult Df (1)Exel6234/Yhs-hid flies in fresh bottles for 24 h to deposit eggs at 22°C. Heat shock incubate eggs at 37°C for 75 min to trigger hid-mediated apoptosis of males and balancer heterozygotes. We recommend using a water bath heated to 37°C for efficient heat shock instead of simply placing the samples in a 37°C incubator. Collect surviving female progeny and rear in groups of 40 for 5–8 days under standard conditions at 25°C to produce age-matched, virgin females. Raising at 29°C can speed up the process of obtaining virgin females. Following heat shock treatment, inspect all eclosed adults to confirm complete absence of male progeny. The Df(1)Exel6234/Yhs-hid system should yield 100% female populations when properly executed.

Chamber specifications:

Structure: Reinforced with four stainless steel columns (3 mm diameter) to ensure stability (Fig. 1D-E and 6A).

Material: Four transparent acrylic plates (180 × 75 mm) with alternating thicknesses (1 cm and 0.3 cm), assembled in a “thick-thin-thin-thick” sequence (Fig. 1F and 6C).

Layout: 36 individual chambers arranged in 3×4 clusters (12 chambers per cluster).

Chamber spacing: 1 mm between individual chambers; 2 cm between clusters. Positioned 3 cm from the chamber’s long edge and 1 cm from the short edge (Fig. 1E-F).

Chamber dimensions are not fixed. By simply adjusting the magnification of the recording lens, users can re-scale the field of view and then refine the x (horizontal) and y (vertical) grid values in the DrosoMating interface to enclose the entire fly activity area within the corresponding wells. After this one-step calibration, the pipeline quantifies mating metrics (CL, CI, MD, etc.) with identical accuracy across different chamber sizes.

Software implementation: Automated analysis is performed using the DrosoMating pipeline (GitHub https://github.com/hcls-kimlab/DrosoMating) (Liu, 2024), the raw code is under win-ff module. which extracts courtship and mating metrics from video recordings. (Linux version: https://github.com/liuxiao916/Fruit_Flies)

Setup of mating assay chamber apparatus and architecture of workstation.

(A) Disassembly of the chamber apparatus. Each chamber comprises a roof, two activity areas, and a floor integrated with four pillars. (B) Workstation for assembling mating assay chambers. Essential required equipments for chamber assembly are shown: a stereomicroscope with an integrated light source, carbon dioxide anesthesia apparatus, a soft brush with feather on the opposite end, chamber apparatus, and fly containing vials. (C) 3D schematic of the mating chamber assembly. Assembly from bottom to top as shown. Briefly, first assemble the floor with one activity area 2 for placing females, then cover the females and separate activity area 1 from activity area 2 using the sex-separation slider. Next, position males and secure the chambers with paper tape. (D) The sex-separation slider is positioned between activity area 1 and activity area 2. (E) The mating chamber assembled with paper tape; the sex-separation slider should be marked with a pen to prevent forgetting to remove it. (F) 3D schematic of single chamber unit, can hold up to 36 pairs of fruit flies simultaneously. (G) Multiple chambers are set up with four chambers vertically aligned on an LED backlight, allowing the accommodation of 144 pairs of fruit flies across all chambers. (H) Video recording setup. Position a 1080p resolution camera or a smartphone with matching video capabilities above the chambers, turn on the LED backlight, and remove sex-separation slider before recording. (I) Setup for recording mating behavior using a camera mounted on a stand, positioned vertically above chambers arranged on a size-matched LED backlight to ensure uniform illumination. The camera is aligned to capture the entire chamber array, enabling simultaneous recording of mating interactions under controlled lighting conditions.

Mating Behavior Assay Step-by-step Procedures

Fly preparation: On day 5 post-eclosion, select healthy males (of the desired genotype) and virgin females. Lightly anesthetize flies using ice or CO2 pad (avoid prolonged exposure to prevent stress) (Fig. 6B). To minimize behavioral perturbations, flies should only be anesthetized at this step. For brief immobilization (e.g., sorting or transferring individuals), use CO2 anesthesia (≤30 s exposure). To avoid bias, all experiments should be done blindly. Label fly tubes with random numbers instead of genotypes and keep a separate list matching numbers to genotypes. Only check which numbers correspond to which genotypes after finishing all data analysis.

Chamber assembly and fly loading: Using a soft brush, gently place a single female into a chamber hole in Floor (Activity Area 2) (Fig. 6B). Cover the chamber with a transparent film to prevent physical contact while allowing visual/chemical interactions. Align Activity Area 1 on top of the film (Fig. 6C). Introduce a single male into the opposite side of the film using the transparent film as a slider to minimize CO2 exposure (Fig. 6D). Limit male exposure to CO2 during this step to avoid behavioral artifacts. Seal the chamber securely with Roof using disposable paper tape (Fig. 6E).

Recovery and Mating Initiation: Transfer the assembled chamber to a 25°C incubator (humidity controlled at 60–70% RH) for 90 min to ensure flies recover fully from handling stress. After recovery, gently remove the transparent film to initiate physical mating (Fig. 6E and F).

Video Recording Setup: Position an adjustable LED panel beneath the chamber to ensure uniform illumination (Fig. 6G). Record mating behavior at 1080p resolution for 1.5–2 h (adjust duration as needed) (Fig. 6H). A minimal setup for capturing fruit fly mating behaviors can be rapidly established without extensive external facility requirements. The essential components include a camera stand, an LED light panel, and a camera capable of recording at a resolution and frame rate of 1080p (30 fps). (Fig. 6I).

Analysis of Drosophila Mating Video Data Analysis and Statistics

The use of the Windows version of DrosoMating

Prepare computer equipped either Windows (or Linux) PC with DrosoMating (ff-chose.exe) installed. The original code instructions for use are provided under the win-ff module.

Transfer video files into the computer to be analyzed. Launch DrosoMating and open the video file.

Enter the number of boards and velocity parameter. After entering the information, click “Enter” to proceed to the next step. (Fig. 1A)

Select the four steel marker pillars in clockwise order (Fig. 1A). Adjust the selection sequence based on the video orientation. For each plate, a preview image will appear. Click on three distinct flies to enhance tracking accuracy (Fig. 1B).

After selecting all regions, return to the DrosoMating home interface. Adjust x (horizontal) and y (vertical) values to align grids with chamber holes. Fine-tune fly detection by modifying the s value (sensitivity threshold) until all flies are correctly identified. Use the grid to segment individual flies (as shown in Fig. 1C).

Note on the s value: The parameter s represents the grayscale intensity threshold (range: 0–255 for 8-bit images) used for binary segmentation of flies from the background. It is automatically calculated as the maximum grayscale value at the three reference fly positions plus an offset of 28, and can be manually adjusted to accommodate varying illumination conditions.

Run the analysis (Save the processed video but not recommended for routine use).

Export raw data as a CSV or Excel file. Open it in spreadsheet software for further processing.

The usage instructions for the Linux version of DrosoMating are presented in Fig. S2

Statistical Analysis

To ensure robust statistical analysis, each experimental group included at least 100 male flies (naïve, sexually experienced, or singly reared). Internal controls were incorporated into every experiment as recommended by Bretman et al. (2011) (Bretman et al., 2011). Normality of the mating duration data was confirmed using the Kolmogorov-Smirnov test (p>0.05). For group comparisons, two-sided Student’s t-tests were applied to calculate significance levels (****p<0.0001, ***p<0.001, **p<0.01, * p<0.05), comparisons among three or more groups were performed using one-way ANOVA with Tukey’s HSD post-hoc tests. Two-factor experimental comparisons were performed using two-way ANOVA followed by Sidak’s post-hoc test. Estimation statistics (Claridge-Chang and Assam, 2016) were additionally used to visualize effect sizes, mean differences, and precision, avoiding reliance solely on null hypothesis testing. All analyses, including data plotting, were performed using GraphPad Prism software.

Discussion

Advancing Courtship Behavior Analysis in Drosophila: From Manual Tracking to computer-vision classifier

The study of male Drosophila mating behavior represents a cornerstone in neurogenetics, providing critical insights into the neural and molecular mechanisms underlying innate social behaviors (Dauwalder, 2008; Dukas, 2020). Historically, the CI—quantifying the proportion of time spent in courtship activities—has served as a fundamental metric for evaluating mating activity (Greenspan and Ferveur, 2000a; Griffith and Ejima, 2009b; Pavlou and Goodwin, 2013; Sokolowski, 2001; Yamamoto and Koganezawa, 2013). However, traditional methodologies relied on labor-intensive manual observation, limiting their capacity to capture dynamic behavioral repertoires such as orientation, wing extension (“courtship song”), and copulation sequences (Dauwalder, 2008; Dukas, 2020). While machine vision systems like Dankert et al.’s automated platform advanced quantification of social interactions, their primary application skewed toward male-male aggression rather than nuanced courtship dynamics (Dankert et al., 2009b). Concurrently, research into genetic and environmental modulators—including fruitless (fru) and doublesex (dsx) gene pathways, cuticular hydrocarbons, and sensory cues—highlighted the need for scalable, high-resolution tools to dissect multimodal influences on behavior.

Technical Innovation and Validation

DrosoMating addresses these gaps through an integrated framework combining modular environmental chambers, automated video tracking, and machine-based annotation. This system minimizes manual intervention while capturing temporally precise metrics such as CL, CI, and MD. Validation against ground-truth manual scoring demonstrated exceptional accuracy (98–99% agreement; error rates <0.05%), underscoring reliability for high-throughput screens. Crucially, the platform detected subtle behavioral variations across genotypes—e.g., diminished courtship vigor in w1118 and y1 mutants—and environmental manipulations such as group-rearing or prior sexual experience, which respectively shortened mating duration by ∼8–12 %. This sensitivity enables investigations into adaptive trade-offs, learning phenotypes, and circuit-level mechanisms. For instance, DrosoMating replicated classic deficits in aversive learning following mushroom body ablation and courtship impairments in rutabaga learning mutants, corroborating its utility for behavioral neuroscience (Levin et al., 1992).

Limitations and Future Directions

Despite its advantages, DrosoMating currently lacks granularity for posture-specific interactions (e.g., “wing threat” during aggression or “circling” during courtship) and multisensory integration. Future iterations could incorporate 3D pose estimation algorithms (e.g., DeepLabCut) or integrate JAABA-based classifiers to expand action repertoires (Kabra et al., 2013). Although DrosoMating does not currently export the full set of per-frame kinematic features required for direct JAABA classifier training, its modular video-processing framework could be extended in future versions to generate JAABA-compatible trajectory or feature outputs.

Cross-species testing was not performed in this study. Although DrosoMating’s state-detection approach is morphology-agnostic and therefore theoretically applicable across Drosophila species, we have revised the text to avoid claiming demonstrated cross-species performance. Formal validation across diverse species remains a promising future direction.

Comparative Evaluation with Existing Courtship Analysis Tools

A major advantage of DrosoMating is its compatibility with low-quality, high-throughput mating videos. Conventional tracking-based tools such as Ctrax and FlyTracker are powerful for trajectory- and pose-based behavioral analysis, but they generally require stable object segmentation, identity maintenance, and reliable feature extraction across frames. Ctrax was designed to estimate the positions and orientations of multiple walking flies while maintaining their identities, whereas FlyTracker aims to track detailed fly pose and generate per-frame behavioral features. Under our recording conditions, these requirements were difficult to satisfy because the videos were low contrast and the male and female frequently overlapped during copulation.

In our tests, both tools showed limited compatibility with these videos. Even after cropping to single-chamber videos and adjusting detection parameters, Ctrax produced unstable target numbers, fragmented trajectories, and tracking errors. FlyTracker could generate a background model and occasionally segment flies successfully, but detection remained unstable and the full pipeline failed during feature computation. These issues are particularly relevant for copulation latency and mating duration analysis: during copulation, the male and female remain physically coupled for a long period, which makes identity-based tracking difficult. For these timing metrics, it is more important to robustly detect the onset and offset of the mating state than to reconstruct detailed individual trajectories. DrosoMating was purpose-built to address these specific challenges. It operates reliably on lower-quality video streams, requires no complex pre-processing or manual ROI definition, and is optimized for high-throughput multi-chamber analysis. While it does not offer the same level of pose or kinematic detail as other tools, it provides a unique solution for laboratories seeking a simple, fast, and robust pipeline to quantify core reproductive timing metrics—copulation latency (CL), courtship index (CI), and mating duration (MD)—without the overhead of more complex systems (Table.1).

Although we directly tested only Ctrax and FlyTracker, this limitation may also affect workflows that depend on upstream tracking-derived features. For example, JAABA uses tracking-derived features to train behavior classifiers, and DANCE, a recent Drosophila aggression and courtship pipeline, uses JAABA-based classifiers and lists FlyTracker and JAABA as required software. Thus, our conclusion is not that these tools are generally unsuitable for Drosophila behavior analysis, but that DrosoMating provides a more practical workflow for low-quality, high-throughput videos focused specifically on mating timing.

While DrosoMating currently prioritizes mating timing metrics over discrete behavioral classification, its modular architecture provides a foundation for future integration with behavior classifiers such as JAABA. Laboratories requiring granular behavioral elements—such as wing extension or circling—would benefit from an extended pipeline that exports per-frame kinematic features for downstream classifier training. Validating this integration represents a promising future direction to broaden the tool’s utility beyond core reproductive timing assays.